---
title: "Census Exploration"
output: html_document
date: "2025-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(sf)
library(dplyr)
library(ggplot2)
library(data.table)
library(tidycensus)
library(tigris)
library(tidyverse)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(scales)
library(leaflet)


# Set up theme and options
colors <- c("0" = "#e42524", "1" = "#008d8a")

map_theme <- theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

options(tigris_use_cache = TRUE)
options(scipen = 999)
```

# Load base datasets
```{r}
# 2022 5-year estimates (2018-2022)
census_tracts <- st_read("C:/Users/rache/Documents/Github/Philly-Homeowners/Data/phila_census1.gpkg")
```


# Properties
Zoning Codes used:
RSA1, RSA2, RSA3, RSA4, RSA5 (Residential Single-Family Attached)
RM1, RM2, RM3, RM4 (Residential Multi-Family)
CMX1, CMX2, CMX2.5, CMX3, CMX4, CMX5 (Commercial Mixed-Use)
IRMX (Industrial Residential Mixed-Use)


Added 2/11/25
RSD-1, RSD-2, RSD-3 (Residential Single-Family Detached)
RTA-1 (Residential Two-Family Attached)
RMX-1, RMX-2, RMX-3 (Residential Mixed-Use)
```{r}
# Philadelphia property data
properties <- fread("C:/Users/rache/Documents/opa_properties_public.csv")


filtered_properties <- properties %>%
  mutate(exemption = ifelse(homestead_exemption == 0, 0, 1))

filtered_properties <- filtered_properties %>% 
  mutate(is_residential = ifelse(zoning %in% c(
    # Single Family Detached
    "RSD1", "RSD2", "RSD3",
    
    # Single Family Attached
    "RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RSA6",
    
    # Two-Family Attached
    "RTA1",
    
    # Multi-Family
    "RM1", "RM2", "RM3", "RM4",
    
    # Residential Mixed-Use
    "RMX1", "RMX2", "RMX3",
    
    # Commercial Mixed-Use (allows residential)
    "CMX1", "CMX2", "CMX2.5", "CMX3", "CMX4", "CMX5",
    
    # Industrial Residential Mixed-Use
    "IRMX"), 1, 0))


# Create indicator for properties where mailing address matches property location
# This helps identify owner-occupied properties, though not perfectly since:
# - Some owners use PO boxes or different mailing addresses
# - Some property management companies use property address for mail
filtered_properties <- filtered_properties %>% 
  mutate(same_address = ifelse(mailing_street == location, 1, 0))

# Create indicator for potential "loophole" properties that have building tax exemptions
# but no homestead exemption, which could indicate:
# - Non-profit owned residential properties
# - Properties with other types of tax abatements
# - Properties that may be eligible for but not utilizing homestead exemption
filtered_properties <- filtered_properties %>%
  mutate(likely_loop = ifelse(exempt_building > 0 & exemption == 0, 1, 0))

# Flag unusually deep lots (over 300 feet)
# Deep lots may indicate:
# - Multiple buildings on one lot
# - Irregular property configurations
# - Potential subdivision candidates
filtered_properties <- filtered_properties %>%
  mutate(is_deep = ifelse(depth > 300, 1, 0))


#names(filtered_properties)

```


```{r}
# Convert to spatial and join with census tracts
properties_sf <- st_as_sf(filtered_properties, wkt = "shape", crs = 2272)
census_tracts <- st_transform(census_tracts, 2272)
properties_tract <- st_join(properties_sf, census_tracts)

# Create tract summary
tract_summary <- properties_tract %>%
  group_by(GEOID) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100,
    .groups = "drop"
  ) %>%
  st_drop_geometry()

# Create final enriched dataset
census_tracts_enriched <- census_tracts %>%
  left_join(tract_summary, by = "GEOID")

```

```{r}
# Transform data to WGS84 (required for leaflet)
census_tracts_wgs84 <- st_transform(census_tracts_enriched, 4326)

# Create interactive map
leaflet(census_tracts_wgs84) %>%
  addTiles() %>%  # Add OpenStreetMap base map
  addPolygons(
    fillColor = ~colorNumeric(
      palette = "viridis",
      domain = c(0, 100)
    )(pct_homestead),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    popup = ~paste(
      "Census Tract:", GEOID, "<br>",
      "Homestead %:", round(pct_homestead, 1), "<br>",
      "Population Density:", round(pop_density, 0), "<br>",
      "Total Properties:", total_properties, "<br>",
      "Median Income:", scales::dollar(median_income)
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("viridis", domain = c(0, 100)),
    values = ~pct_homestead,
    title = "% Homestead Exemption",
    opacity = 0.7
  )

```
```{r}
census_tracts_filtered <- census_tracts_enriched %>%
  filter(pop_density > 0 & total_properties >= 30)  # 30 properties minimum

census_tracts_invalid <- census_tracts_enriched %>%
  filter(pop_density == 0 | total_properties < 30)  # Include low property counts in "invalid"


# Transform both to WGS84
census_tracts_invalid_wgs84 <- st_transform(census_tracts_invalid, 4326)
census_tracts_filtered_wgs84 <- st_transform(census_tracts_filtered, 4326)

# Create map with both layers
leaflet() %>%
  addTiles() %>%
  # Add invalid tracts first (in gray)
  addPolygons(data = census_tracts_invalid_wgs84,
    fillColor = "gray",
    fillOpacity = 0.5,
    weight = 1,
    color = "white",
    popup = "No data available"
  ) %>%
  # Add valid tracts with your original styling
  addPolygons(data = census_tracts_filtered_wgs84,
    fillColor = ~colorNumeric(
      palette = "viridis",
      domain = c(0, 85)
    )(pct_homestead),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    popup = ~paste(
    "Census Tract:", GEOID, "<br>",
    "Homestead %:", round(pct_homestead, 1), "%<br>",
    "Total Properties:", total_properties,
    ifelse(total_properties < 100, 
           "<br><i style='color:red'>Note: Low property count may affect reliability</i>", 
           "")
  )

  )%>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("viridis", domain = c(0, 85)),
    values = census_tracts_filtered_wgs84$pct_homestead,
    title = "% Homestead Exemption",
    opacity = 0.7,
    labFormat = labelFormat(suffix = "%")
  ) %>%
  # Add legend for gray areas
  addLegend(
    position = "bottomright",
    colors = "gray",
    labels = "No Data Available",
    opacity = 0.5
  )


```


```{r}
# Create histogram of homestead exemption distribution
census_hist <- ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), # Filter out zero population tracts
       aes(x = pct_homestead)) +
  geom_histogram(
    binwidth = 5,
    fill = "#008d8a",
    color = "white"
  ) +
  labs(
    title = "Distribution of Homestead Exemption Rates Across Philadelphia Census Tracts",
    subtitle = "Excluding Zero Population Density Tracts",
    x = "Percentage of Properties with Homestead Exemption",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14)
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10))

census_hist
census_hist <-
#ggsave("outputs/homestead-exemption-distribution.png", census_hist, width = 10, height = 6)
```
The distribution shows a roughly normal shape with most tracts clustered between 30-50%
There's a notable drop-off below 30% in the number of tracts
The histogram shows relatively few tracts with rates below 20%

Therefore, census tracts with homestead exemption rates below 30% could be considered to have low enrollment and might warrant targeted outreach or investigation into barriers to participation, assuming they are primarily residential areas and not institutional/special use tracts.

There's a notable drop-off below 30% in the number of tracts. Therefore, census tracts with homestead exemption rates below 30% could be considered to have low enrollment and might warrant targeted outreach or investigation into barriers to participation, assuming they are primarily residential areas.

```{r}
homestead_pattern <- ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = cut(pct_homestead, 
              breaks = c(0, 20, 30, 40, 50, 60, 100),
              labels = c("<20%", "20-30%", "30-40%", "40-50%", "50-60%", ">60%")))) +
  scale_fill_viridis_d(
    name = "Homestead\nExemption Rate",
    na.value = "gray80",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    title = "Homestead Exemption Rates Across Philadelphia",
    subtitle = "By Census Tract (Excluding Zero Population Areas)",
    caption = "Gray areas indicate zero population density tracts"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_blank(),
    legend.text = element_text(size = 14)
  )
```


```{r}
homestead_pattern
#ggsave("outputs/homestead-exemption-pattern.png", homestead_pattern, width = 10, height = 6)
```
```{r}
# Basic summary statistics
summary(census_tracts_enriched$pop_density)

# More detailed statistics
census_tracts_enriched %>%
  summarise(
    mean_density = mean(pop_density, na.rm = TRUE),
    median_density = median(pop_density, na.rm = TRUE),
    q1 = quantile(pop_density, 0.25, na.rm = TRUE),
    q3 = quantile(pop_density, 0.75, na.rm = TRUE)
  )

# Visual distribution
ggplot(census_tracts_enriched, aes(x = pop_density)) +
  geom_histogram(binwidth = 1000) +
  theme_minimal() +
  labs(title = "Distribution of Population Density in Philadelphia Census Tracts",
       x = "Population Density (per square mile)",
       y = "Count of Census Tracts")

```
```{r}
ggplot(census_tracts_enriched, aes(x = pop_density)) +
  geom_histogram(binwidth = 1000) +
  theme_minimal() +
  labs(title = "Distribution of Population Density in Philadelphia Census Tracts",
       x = "Population Density (per square mile)",
       y = "Count of Census Tracts")
```


```{r}
homestead_pattern <- ggplot(census_tracts_enriched %>% 
                           mutate(pct_homestead = case_when(
                             pop_density < 2000 | total_properties < 100 ~ NA_real_,
                             TRUE ~ pct_homestead))) +
  geom_sf(aes(fill = cut(pct_homestead, 
              breaks = c(0, 20, 30, 40, 50, 60, 100),
              labels = c("<20%", "20-30%", "30-40%", "40-50%", "50-60%", ">60%")))) +
  scale_fill_viridis_d(
    name = "Homestead\nExemption Rate",
    na.value = "gray80",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    title = "Homestead Exemption Rates Across Philadelphia",
    subtitle = "By Census Tract",
    caption = "Gray tracts indicate population density < 2,000 per sq. mile or fewer than 100 properties"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_blank(),
    legend.text = element_text(size = 14)
  )

```

```{r}
homestead_pattern
```



```{r}
# Distribution of pct_homestead
summary(tract_summary$pct_homestead)

# Map for low rates (under 30%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density > 0 & pct_homestead < 30, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Under 30%)",
    na.value = "gray80",
    limits = c(0, 30),
    breaks = seq(0, 30, by = 5)
  ) +
  labs(
    title = "Low Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Below 30% Enrollment",
    caption = "Gray areas: Zero population density or rates ≥ 30%"
  ) +
  map_theme

# Table for low rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density > 0 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

# Map for high rates (over 60%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density > 0 & pct_homestead > 60, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Over 60%)",
    na.value = "gray80",
    limits = c(60, 82),
    breaks = seq(60, 80, by = 5)
  ) +
  labs(
    title = "High Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Above 60% Enrollment",
    caption = "Gray areas: Zero population density or rates ≤ 60%"
  ) +
  map_theme

# Table for high rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density > 0 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

```

```{r}
# Distribution of pct_homestead with new criteria
summary(tract_summary$pct_homestead[tract_summary$pop_density >= 2000 & tract_summary$total_properties >= 100])

# Map for low rates (under 30%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density >= 2000 & total_properties >= 100 & pct_homestead < 30, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Under 30%)",
    na.value = "gray80",
    limits = c(0, 30),
    breaks = seq(0, 30, by = 5)
  ) +
  labs(
    title = "Low Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Below 30% Enrollment",
    caption = "Gray areas: Low density (<2000/sq mi), low property count (<100), or rates ≥ 30%"
  ) +
  map_theme

# Table for low rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & total_properties >= 100 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

# Map for high rates (over 60%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density >= 2000 & total_properties >= 100 & pct_homestead > 60, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Over 60%)",
    na.value = "gray80",
    limits = c(60, 82),
    breaks = seq(60, 80, by = 5)
  ) +
  labs(
    title = "High Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Above 60% Enrollment",
    caption = "Gray areas: Low density (<2000/sq mi), low property count (<100), or rates ≤ 60%"
  ) +
  map_theme

# Table for high rates
high_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & total_properties >= 100 & pct_homestead > 60) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(desc(pct_homestead))

high_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(desc(pct_homestead)) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with High Homestead Exemption Rates (>60%)"
  )

```

```{r}
census_tracts_enriched <- census_tracts_enriched %>%
  mutate(
    owner_occ_rate = (census_tracts_enriched$owner_hh / census_tracts_enriched$occupied_units) * 100
  )
    
    
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & 
         total_properties >= 100 & 
         pct_homestead < 30 &
         owner_occ_rate > 40) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(pct_homestead)

# For the map visualization
low_enrollment_tracts_map <- ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = case_when(
    pop_density >= 2000 & 
    total_properties >= 100 & 
    pct_homestead < 30 &
    owner_occ_rate > 40 ~ "#f4aa9e",
    TRUE ~ "gray80"
  ))) +
  scale_fill_identity() +
  labs(
    title = "Low Homestead Exemption Tracts",
    subtitle = "Tracts with <30% Homestead Rate & >40% Owner Occupancy",
    caption = "Gray areas do not meet filtering criteria"
  ) +
  map_theme




low_enrollment_tracts_map
#ggsave("outputs/homestead-exemption-low_enrollment_tracts.png", low_enrollment_tracts_map, width = 10, height = 6)
```
```{r}
# For the table
low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%) and High Owner Occupancy (>40%)"
  )
```


```{r}
high_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & 
         total_properties >= 100 & 
         pct_homestead < 60 &
         owner_occ_rate > 40) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(pct_homestead)

# For the map visualization
low_enrollment_tracts_map <- ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = case_when(
    pop_density >= 2000 & 
    total_properties >= 100 & 
    pct_homestead < 60 &
    owner_occ_rate > 40 ~ "#f4aa9e",
    TRUE ~ "gray80"
  ))) +
  scale_fill_identity() +
  labs(
    title = "Low Homestead Exemption Tracts",
    subtitle = "Tracts with <60% Homestead Rate & >40% Owner Occupancy",
    caption = "Gray areas do not meet filtering criteria"
  ) +
  map_theme




low_enrollment_tracts_map
#ggsave("outputs/homestead-exemption-high_enrollment_tracts.png", high_enrollment_tracts_map, width = 10, height = 6)
```


```{r}
ggplot(census_tracts_enriched %>% 
       filter(pop_density >= 2000), # Using your established density threshold
       aes(x = owner_occ_rate)) +
  geom_histogram(
    binwidth = 5,
    fill = "#e42524",  # Using your established red color
    alpha = 0.8,
    color = "white"
  ) +
  labs(
    title = "Distribution of Owner Occupancy Rates Across Philadelphia Census Tracts",
    subtitle = "Excluding Low Density Areas (<2,000 per square mile)",
    x = "Owner Occupancy Rate (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10))

```



```{r}
# Scatterplot of homestead rates vs median home values
ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), 
       aes(x = median_home_value, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  labs(
    title = "Homestead Exemption Rates vs. Home Values",
    x = "Median Home Value",
    y = "Percentage with Homestead Exemption"
  )

#use liner model =lm
```

```{r}
# Create the linear model
homestead_model <- lm(pct_homestead ~ median_home_value, 
                     data = census_tracts_enriched %>% 
                     filter(pop_density > 0))

# View the summary statistics
summary(homestead_model)

# Visualize with linear fit instead of loess
ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), 
       aes(x = median_home_value, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +  # Changed from loess to lm
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  labs(
    title = "Linear Relationship: Homestead Exemption Rates vs. Home Values",
    x = "Median Home Value",
    y = "Percentage with Homestead Exemption"
  )

```

The scatterplot shows the relationship between median home values (x-axis) and homestead exemption rates (y-axis) across Philadelphia census tracts. The pattern suggests:

Homestead exemption rates increase with home values up to around $250,000. 

Peak participation occurs in the $200,000-$300,000 range (around 50%).

There's a slight decline in participation for higher-value homes.

The widening gray area at higher home values indicates more uncertainty in the trend, likely due to fewer data points in that range.

Wide variation in participation rates across all home values, shown by the vertical spread of points

```{r}
# Owner Occupancy vs Homestead Rates Analysis
occupancy_analysis <- census_tracts_enriched %>%
  filter(pop_density > 0) %>%
  mutate(
    owner_occ_rate = (owner_hh / occupied_units) * 100,
    pct_homestead_owners = (homestead_count / owner_hh) * 100
  ) %>%
  select(GEOID, owner_occ_rate, pct_homestead_owners)

# Visualize the relationship
ggplot(occupancy_analysis, aes(x = owner_occ_rate, y = pct_homestead_owners)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(
    title = "Owner Occupancy Rate vs. Homestead Participation",
    x = "Owner Occupancy Rate (%)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()

ggsave("outputs/homestead-exemption-distribution.png", census_hist, width = 10, height = 6)
```
This scatter plot reveals important patterns in homestead exemption participation across Philadelphia's neighborhoods. By comparing census tract data on owner occupancy rates (from the 2022 5-year ACS) with homestead exemption enrollment, we can identify areas where participation could be improved.

The data shows that while most Philadelphia census tracts have owner occupancy rates between 25-75%, and generally over half of eligible homeowners participate in the program, there are clear opportunities for improvement. Particularly concerning are:

Census tracts with participation rates below 50%

Areas with high owner occupancy but low program participation

Neighborhoods falling well below the expected participation rate (shown by the blue trend line)

Notably, higher rates of owner occupancy don't automatically translate to higher program participation. This suggests that other factors beyond home ownership - such as awareness of the program, ease of enrollment, or demographic characteristics - may play more significant roles in determining participation rates. These insights can help guide targeted outreach efforts to increase program enrollment among eligible homeowners who are currently missing out on this tax benefit.

```{r}
# Create the dataset with the calculated rates
occupancy_analysis <- census_tracts_enriched %>%
  filter(pop_density > 0) %>%
  mutate(
    owner_occ_rate = (owner_hh / occupied_units) * 100,
    pct_homestead_owners = (homestead_count / owner_hh) * 100
  ) %>%
  select(GEOID, owner_occ_rate, pct_homestead_owners)

# Fit the linear model
occupancy_model <- lm(pct_homestead_owners ~ owner_occ_rate, data = occupancy_analysis)

# View model summary
summary(occupancy_model)

# Visualize with linear fit
ggplot(occupancy_analysis, aes(x = owner_occ_rate, y = pct_homestead_owners)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +  # Changed from loess to lm
  labs(
    title = "Linear Relationship: Owner Occupancy vs. Homestead Participation",
    x = "Owner Occupancy Rate (%)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()

```


# Zoning Analysis
```{r}
# Data frame of zoning types
zoning_types <- data.frame(
  Type = c(
    "Single Family Detached",
    "Single Family Attached",
    "Two-Family Attached",
    "Multi-Family",
    "Residential Mixed-Use",
    "Commercial Mixed-Use",
    "Industrial Residential Mixed-Use"
  ),
  Codes = c(
    "RSD1, RSD2, RSD3",
    "RSA1, RSA2, RSA3, RSA4, RSA5, RSA6",
    "RTA1",
    "RM1, RM2, RM3, RM4",
    "RMX1, RMX2, RMX3",
    "CMX1, CMX2, CMX2.5, CMX3, CMX4, CMX5",
    "IRMX"
  ),
  Description = c(
    "Detached houses on individual lots",
    "Attached and semi-detached houses on individual lots",
    "Two-family, semi-detached houses on individual lots",
    "Moderate to high-density multi-unit residential buildings",
    "Residential and mixed-use development, including master plan development",
    "Neighborhood to regional-serving mixed-use development",
    "Mix of low-impact industrial, artisan industrial, residential, and neighborhood commercial uses"
  )
)

# Formatted table
kable(zoning_types,
      col.names = c("Residential Type", "Zoning Codes", "Description"),
      caption = "Philadelphia Residential Zoning Classifications")

```


```{r}
properties_filtered <- filtered_properties %>%
  select(
    zoning,
    homestead_exemption,
    is_residential,
    census_tract,
    shape
  )

# First create a zoning type classification
filtered_properties <- filtered_properties %>% 
  mutate(
    zoning_type = case_when(
      zoning %in% c("RSD1", "RSD2", "RSD3") ~ "Single Family Detached",
      zoning %in% c("RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RSA6") ~ "Single Family Attached",
      zoning %in% c("RTA1") ~ "Two-Family Attached",
      zoning %in% c("RM1", "RM2", "RM3", "RM4") ~ "Multi-Family",
      zoning %in% c("RMX1", "RMX2", "RMX3") ~ "Residential Mixed-Use",
      zoning %in% c("CMX1", "CMX2", "CMX2.5", "CMX3", "CMX4", "CMX5") ~ "Commercial Mixed-Use",
      zoning %in% c("IRMX") ~ "Industrial Residential Mixed-Use",
      TRUE ~ "Other"
    ),
    is_residential = ifelse(zoning_type != "Other", 1, 0)
  )


zoning_summary <- filtered_properties %>%
  filter(is_residential == 1) %>%
  group_by(zoning_type) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100
  ) %>%
  arrange(desc(pct_homestead))


# Homestead rates by zoning type
zoning_summary_chart <- ggplot(zoning_summary, 
       aes(x = reorder(zoning_type, -pct_homestead), 
           y = pct_homestead)) +
  geom_bar(stat = "identity", 
           fill = "#e42524",
           alpha = 0.8) +
  geom_text(aes(label = round(pct_homestead,1)), 
            vjust = -0.5,
            size = 3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14)
  ) +
  labs(
    title = "Homestead Exemption Rates by Zoning Category in Philadelphia",
    subtitle = "Residential and Mixed-Use Districts Only",
    x = "Zoning Category",
    y = "Percentage with Homestead Exemption",
    caption = "Source: Philadelphia Property Data, 2025"
  ) +
  scale_y_continuous(
    limits = c(0, max(zoning_summary$pct_homestead) * 1.1),
    labels = function(x) paste0(x, "%")
  )



zoning_summary_chart
ggsave("outputs/homestead-exemption-zoning_summary_chart.png", zoning_summary_chart, width = 10, height = 6)
```


```{r}
assessments <- fread("C:/Users/rache/Documents/assessments.csv") %>%
  filter(year %in% c(2022, 2025))

```
```{r}
# Use more conservative inflation adjustment
inflation_multiplier <- 1.09  # About 3% per year

assessments_change <- assessments %>%
  filter(year %in% c(2022, 2025)) %>%
  filter(market_value > 0) %>%  # Remove zero values
  group_by(parcel_number) %>%
  arrange(year) %>%
  summarize(
    value_2022 = first(market_value),
    value_2022_adjusted = first(market_value) * inflation_multiplier,
    value_2025 = last(market_value),
    real_change = value_2025 - value_2022_adjusted,
    real_percent_change = ((value_2025 - value_2022_adjusted) / value_2022_adjusted) * 100
  ) %>%
  filter(!is.infinite(real_percent_change)) %>%  # Remove infinite values
  ungroup()

```



```{r}
properties_with_assesment_change <- filtered_properties %>%
  left_join(assessments_change, by = "parcel_number")
```


```{r}
# First create the spatial properties with assessment changes
properties_changes_sf <- st_as_sf(properties_with_assesment_change, wkt = "shape", crs = 2272)

# Join with census tracts
properties_changes_tract <- st_join(properties_changes_sf, census_tracts)

# Create tract summary including assessment changes
tract_assessment_summary <- properties_changes_tract %>%
  group_by(GEOID) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100,
    avg_real_percent_change = mean(real_percent_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_drop_geometry()

# Create final enriched dataset with assessment changes
census_tracts_enriched <- census_tracts %>%
  left_join(tract_assessment_summary, by = "GEOID")

```


```{r}
# Transform to WGS84 for mapping
census_tracts_wgs84 <- st_transform(census_tracts_enriched, 4326)

# Create map of real percent changes
assessment_change_map <- ggplot(census_tracts_enriched %>%
                              filter(pop_density >= 2000 & 
                                     total_properties >= 100)) +
  geom_sf(aes(fill = avg_real_percent_change)) +
  scale_fill_viridis_c(
    name = "Average Real\nValue Change (%)",
    na.value = "gray80"
  ) +
  labs(
    title = "Real Property Value Changes in Philadelphia (2022-2025)",
    subtitle = "Adjusted for Inflation, Excluding Low Density Areas",
    caption = "Gray areas indicate insufficient data or low density tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text = element_blank(),
    axis.title = element_blank()
  )

assessment_change_map

```



##### In progress...
# Further Analysis
```{r}
# Special Use Areas Analysis
special_use_tracts <- census_tracts_enriched %>%
  filter(pop_density > 0 & pct_homestead < 20) %>%
  select(GEOID, pct_homestead, pop_density, total_properties)

# Map special use areas
ggplot(census_tracts_enriched) +
  geom_sf() +
  geom_sf(data = special_use_tracts, fill = "red", alpha = 0.5) +
  labs(
    title = "Potential Special Use Areas",
    subtitle = "Census Tracts with <20% Homestead Exemption Rate"
  ) +
  map_theme
```

## Special Use 
Philadelphia International Airport area
Wissahickon Park
Pennypack Park
Northeast Airport
International Airport
Hunting Park
Stadium District
Industrial-Port
Cobbs Creek Park
Industrial - Refinery

- Parks and Recreation Areas (Wissahickon, Pennypack, Hunting Park)
- Transportation Infrastructure (Airports)
- Industrial Zones (Port, Refinery)
- Sports Complexes (Stadium District)


```{r}
# Property Characteristics Comparison
property_comparison <- census_tracts_enriched %>%
  group_by(pct_homestead > 0) %>%
  summarise(
    avg_year_built = mean(year_built, na.rm = TRUE),
    avg_market_value = mean(market_value, na.rm = TRUE),
    avg_sale_price = mean(sale_price, na.rm = TRUE),
    avg_living_area = mean(total_livable_area, na.rm = TRUE)
  ) %>%
  rename(has_homestead = `pct_homestead > 0`)

# Comparison plots
ggplot(census_tracts_enriched, aes(x = factor(pct_homestead > 0), y = market_value)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Market Values: Homestead vs Non-Homestead Properties",
    x = "Has Homestead Exemption",
    y = "Market Value"
  ) +
  theme_minimal()

```

```{r}
# Filter for homestead properties only
homestead_properties <- properties %>%
  filter(homestead_exemption == "YES")

# aggregations for homestead-only properties
homestead_tract_summary <- st_join(homestead_properties, census_tracts) %>%
  group_by(GEOID) %>%
  summarise(
    avg_year_built_homestead = mean(year_built, na.rm = TRUE),
    avg_market_value_homestead = mean(market_value, na.rm = TRUE),
    avg_sale_price_homestead = mean(sale_price, na.rm = TRUE),
    avg_livable_area_homestead = mean(total_livable_area, na.rm = TRUE)
  )

```

# By Zip Code
```{r}
zip_summary <- properties %>%
  group_by(zip_code) %>%
  


```


######

# Commercial Activity Licenses
```{r}
comlicenses <- read.csv("Data/com_act_licenses.csv")

# Clean
comlicenses_clean <- comlicenses %>% 
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  select(where(~ any(!is.na(.)))) %>%
  select(licensenum, licensestatus, legalfirstname, legallastname, legalentityid)

# Filter for active licenses and create full names
comlicenses_licenses <- comlicenses_clean %>%
  filter(licensestatus == "Active") %>%
  mutate(full_name = paste(legallastname, legalfirstname)) %>%
  distinct(full_name, .keep_all = TRUE)
```


```{r}
# Join commercial licenses with properties
properties_comlicenses <- properties_sf %>%
  left_join(comlicenses_licenses %>%
              select(full_name, licensestatus), 
            by = c("owner_1" = "full_name")) %>%
  left_join(comlicenses_licenses %>%
              select(full_name, licensestatus), 
            by = c("owner_2" = "full_name")) %>%
  mutate(com_potential = ifelse(licensestatus.x == "Active" | 
                               licensestatus.y == "Active", TRUE, FALSE)) %>%
  select(-licensestatus.x, -licensestatus.y) %>%
  mutate(com_potential = replace_na(com_potential, FALSE))
```


# Tax balances
```{r}
balances <- read.csv("Data/real_estate_tax_balances_census_tract.csv")

# Create balance summary by census tract
balance_distinct <- balances %>%
  group_by(census_tract) %>% 
  summarise(avg_balance = sum(balance, na.rm = TRUE) / 
                         sum(num_props, na.rm = TRUE)) %>%
  ungroup()

# Convert census_tract to character in balance_distinct
balance_distinct <- balances %>%
  group_by(census_tract) %>% 
  summarise(avg_balance = sum(balance, na.rm = TRUE) / 
                         sum(num_props, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(census_tract = as.character(census_tract))

# Add all data to census tracts
census_tracts_enriched <- census_tracts_enriched %>%
  left_join(balance_distinct, by = c("GEOID" = "census_tract")) %>%
  left_join(properties_comlicenses %>% 
              st_drop_geometry() %>%
              group_by(census_tract) %>%
              mutate(census_tract = as.character(census_tract)) %>%
              summarise(pct_commercial = mean(com_potential) * 100),
            by = c("GEOID" = "census_tract"))

```











