# Install necessary libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(scales)
library(tidycensus)
library(tigris)
library(sf)
library(data.table)
library(corrr)
options(scipen = 999)
root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```

# Main Property Dataset
```{r}
properties <- fread("Data/opa_properties_public.csv")
```

```{r}
properties_sf<-st_read("Data/property_data_geoid.gpkg")

```


```{r}
census_tracts<-st_read("Data/phila_census1.gpkg")
```


```{r}
# Philadelphia property data

filtered_properties <- properties %>%
  mutate(exemption = ifelse(homestead_exemption == 0, 0, 1))

filtered_properties <- filtered_properties %>% 
  mutate(is_residential = ifelse(zoning %in% c(
    # Single Family Detached
    "RSD1", "RSD2", "RSD3",
    
    # Single Family Attached
    "RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RSA6",
    
    # Two-Family Attached
    "RTA1",
    
    # Multi-Family
    "RM1", "RM2", "RM3", "RM4",
    
    # Residential Mixed-Use
    "RMX1", "RMX2", "RMX3",
    
    # Commercial Mixed-Use (allows residential)
    "CMX1", "CMX2", "CMX2.5", "CMX3", "CMX4", "CMX5",
    
    # Industrial Residential Mixed-Use
    "IRMX"), 1, 0))

```


```{r}
# Convert to spatial and join with census tracts
properties_sf <- st_as_sf(filtered_properties, wkt = "shape", crs = 2272)
census_tracts <- st_transform(census_tracts, 2272)
properties_tract <- st_join(properties_sf, census_tracts)

# Create tract summary
tract_summary <- properties_tract %>%
  group_by(GEOID) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100,
    .groups = "drop"
  ) %>%
  st_drop_geometry()

# Create final enriched dataset
census_tracts_enriched <- census_tracts %>%
  left_join(tract_summary, by = "GEOID")

```

# Commercial Activity Licenses Dataset

Eligibility: Not used exclusively for business purposes or as rental units (a percentage is fine) 

##Potential Variable: F/T exclusively for business purposes (Binary) 

F/T - Does the applicant hold a commercial license?  

# business Licenses
```{r}
business_license<-st_read("Data/business_licenses.geojson")
```

```{r}
#table(business_license$licensetype)
#table(business_license$rentalcategory)
```

```{r}
rental_license<-business_license%>%
  filter(licensetype=="Rental",
         #rentalcategory=="Residential Dwellings",
         licensestatus %in% c("Active"))%>%
  select(opa_account_num)%>%
  st_drop_geometry()%>%
  distinct() %>% 
  filter(!is.na(opa_account_num)) %>% 
  mutate(rental_license = 1)  
```


```{r}
properties_rental <- properties_sf %>%
  mutate(parcel_number = as.character(parcel_number))%>%
  left_join(rental_license, by = c("parcel_number" = "opa_account_num"))%>%
  mutate(rental_license = replace_na(rental_license, 0))

```

```{r}
ggplot(properties_rental, aes(x = factor(rental_license), fill = factor(exemption))) +
  geom_bar(position = "dodge") +  
  geom_text(stat = "count", aes(label = ..count.., color = factor(exemption)),  
            position = position_dodge(width = 0.9),  
            vjust = -0.5,  
            size = 3) +  
  scale_fill_manual(values = c("0" = "#e42524", "1" = "#00ADA9"),  
                    labels = c("0" = "No Exemption", "1" = "With Exemption")) +  
  scale_color_manual(values = c("0" = "#e42524", "1" = "#00ADA9")) +  
  labs(title = "Rental Licenses Metrics by Homestead Exemption Status",
       x = "Rental Licenses",
       y = "Count",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(legend.position = "none") 

```


```{r}
commercial_license<-business_license%>%
  filter(licensestatus %in% c("Active"))%>%
  filter(licensetype %in% c(
    "Food Caterer",
    "Food Establishment, Retail Perm Location (Large)",
    "Food Establishment, Retail Permanent Location",
    "Food Manufacturer / Wholesaler",
    "Food Preparing and Serving",
    "Food Preparing and Serving (30+ SEATS)",
    "Motor Vehicle Repair / Retail Mobile Dispensing",
    "Pawn Shop",
    "Precious Metal Dealer",
    "Public Garage / Parking Lot",
    "Residential Property Wholesaler",
    "Tire Dealer",
    "Tow Company",
    "Vacant Commercial Property"
  ))%>%
  select(opa_account_num)%>%
  st_drop_geometry()%>%
  distinct() %>% 
  filter(!is.na(opa_account_num)) %>% 
  mutate(commercial_license = 1)  
```


```{r}
properties_commercial <- properties_sf %>%
  mutate(parcel_number = as.character(parcel_number))%>%
  left_join(commercial_license, by = c("parcel_number" = "opa_account_num"))%>%
  mutate(commercial_license = replace_na(commercial_license, 0))

```

```{r}
ggplot(properties_commercial, aes(x = factor(commercial_license), fill = factor(exemption))) +
  geom_bar(position = "dodge") +  
  geom_text(stat = "count", aes(label = ..count.., color = factor(exemption)),  
            position = position_dodge(width = 0.9),  
            vjust = -0.5,  
            size = 3) +  
  scale_fill_manual(values = c("0" = "#e42524", "1" = "#00ADA9"),  
                    labels = c("0" = "No Exemption", "1" = "With Exemption")) +  
  scale_color_manual(values = c("0" = "#e42524", "1" = "#00ADA9")) +  
  labs(title = "Commercial Licenses Metrics by Homestead Exemption Status",
       x = "Commercial Licenses (Exclude Rental)",
       y = "Count",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(legend.position = "none") 
```

# Tax balances

May have some negative effect in application 

```{r}
balances <- read.csv("Data/real_estate_tax_balances_census_tract.csv")
```

```{r}
balance_sf <- census_tracts_enriched %>% 
  left_join(balances %>% select(census_tract,balance,num_props), 
            by = c("GEOID" = "census_tract")) 
```

```{r}
ggplot(balance_sf) +
  geom_sf(aes(fill = balance), color = "white", size = 0.1) + 
  scale_fill_gradientn(colors = c("#00ADA9","#e3f9f7","#f4aa9e", "#e42524"),
                       limits = range(balance_sf$balance, na.rm = TRUE),
                       breaks = range(balance_sf$balance, na.rm = TRUE)
                       ) + 
  labs(title = "Total Tax Balance by Census Tract",
       fill = "Price") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 6),
    axis.title = element_text(size = 6),
    axis.text = element_blank(),
    legend.text = element_text(size = 6),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```

```{r}
# Visualize the relationship
ggplot(balance_sf, aes(x = balance, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(
    title = "Principle vs. Homestead Participation",
    x = "Total Principle ($)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()
```


#integrate with property by census tract

```{r}
#distinct
balance_distinct <- balances %>%
  mutate(census = as.numeric(substr(census_tract, 6, 9)))%>%
  st_drop_geometry()%>%
  group_by(census) %>% 
  summarise(avg_balance=sum(balance,na.rm=TRUE)/sum(num_props, na.rm = TRUE),
            sum_balance=sum(balance,na.rm=TRUE),
            tax_props=sum(num_props,na.rm=TRUE)) %>%
  ungroup()
```

```{r}
properties_number<-properties_sf%>%
  select(census_tract,exemption,parcel_number)%>%
  mutate(prop=1)%>%
  st_drop_geometry()%>%
  group_by(census_tract)%>%
  summarise(total_props=sum(prop))%>%
  ungroup()

properties_sf_number<-properties_sf%>%
  left_join(properties_number,by="census_tract")

```


```{r}
properties_balance <- properties_sf_number %>%
  select(objectid, census_tract,exemption,parcel_number,total_props)%>%
  left_join(balance_distinct, by = c("census_tract" = "census"))%>%
  mutate(avg_balance = replace(avg_balance, is.na(avg_balance), 0),
         sum_balance = replace(sum_balance, is.na(sum_balance), 0),
         tax_props=replace(tax_props,is.na(tax_props),0))%>%
  mutate(tax_rate=tax_props/total_props)
  
```

```{r}
avg_values <- properties_balance %>%
  st_drop_geometry()%>%
  group_by(exemption) %>%
  summarise(
    'Total Tax Balance (In 100,000)' = mean(sum_balance, na.rm = TRUE)/100000,
    #Tax_Props=mean(tax_props,na.rm=TRUE),
    '% of Properties with Tax Balance'=mean(tax_rate,na.rm=TRUE)*100
  ) %>%
  pivot_longer(cols = -exemption, names_to = "variable", values_to = "mean_value")

ggplot(avg_values, aes(x = variable, y = mean_value, fill = as.factor(exemption))) +
  geom_bar(stat = "identity", width = 0.5, position = position_dodge(width = 0.6)) +
  
  # 添加数值标签，颜色与柱子一致
  geom_text(aes(label = round(mean_value, 4), color = as.factor(exemption)), 
            position = position_dodge(width = 0.6), 
            vjust = -0.5, size = 5) +

  labs(title = "Mean of Tax Variables by Exemption Status",
       x = "Metrics",
       y = "Mean Value",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  
  # 颜色匹配
  scale_fill_manual(values = c("#e42524", "#00ADA9"), labels = c("No Exemption", "With Exemption")) +
  scale_color_manual(values = c("#e42524", "#00ADA9"), guide = "none") 
```
