---
title: "practicum_Wenjun"
author: "Wenjun Zhu"
date: "2025-03-03"
output: html_document
---

# Install necessary libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(scales)
options(scipen = 999)
root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```

# 1. Data Preparation
# 1.1 Split into Residential and Commercial Properties
```{r}
# Load and inspect the properties dataset
properties <- read.csv("/Users/zhuzhu/Desktop/Penn/25spring/practicum/data/opa_properties_public.csv")
tail(properties, 10)
colnames(properties)
# Add exemption and residential status columns
filtered_properties <- properties %>%
  mutate(exemption = ifelse(homestead_exemption == 0, 0, 1))
filtered_properties <- filtered_properties %>% mutate(is_residential = ifelse(zoning %in% c("RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RM1", "RM1|RSA5", "RM2", "RM3", "RM4","CMX1", "CMX2", "CMX2.5","CMX3", "CMX4", "CMX5","IRMX"), 1, 0))
# Check column names of the filtered dataset
colnames(filtered_properties)
```

# 1.2 Cleaned_properties: Select Relevant Columns
Select key columns for analysis: parcel_number, exemption, is_residential, and shape
```{r}
cleaned_properties <- filtered_properties %>%
  select(parcel_number, exemption, is_residential, shape)

head(cleaned_properties)
colnames(cleaned_properties)
```


# 1.3 Combine Properties Data with Assessment Data
```{r}
# Load and inspect the assessments dataset
assessments2 <- read.csv("~/Desktop/Penn/25spring/practicum/data/assessments.csv")
colnames(assessments2)
head(assessments2, 10)

# Merge properties and assessments data by parcel_number
assessment_combined <- assessments2 %>%
  left_join(cleaned_properties, by = "parcel_number")

head(assessment_combined)
```
# 2. Residential Property History Analysis
## 2.1 Filter Residential Data
```{r}
# Filter the combined dataset to include only residential properties
residential_assessment_combined <- assessment_combined %>%
  filter(is_residential == 1)
head(residential_assessment_combined)
colnames(residential_assessment_combined)
```


## 2.2 Market Value Growth Rate by Exemption Status (2016-2025)
```{r}
library(dplyr)
library(ggplot2)
library(scales)

# Calculate yearly market value and growth rate by exemption status
yearly_market_value <- residential_assessment_combined %>%
  group_by(year, exemption) %>%
  summarise(total_market_value = sum(market_value, na.rm = TRUE), .groups = 'drop')


yearly_market_value <- yearly_market_value %>%
  arrange(exemption, year) %>%
  group_by(exemption) %>%
  mutate(market_value_growth_rate = (total_market_value - lag(total_market_value)) / lag(total_market_value) * 100)

# Filter data for the years 2016-2025
yearly_market_value_filtered <- yearly_market_value %>%
  filter(year >= 2016 & year <= 2025)

# Plot the market value growth rate
ggplot(yearly_market_value_filtered, aes(x = year, y = market_value_growth_rate, color = as.factor(exemption))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, by = 1)) + 
  scale_y_continuous(labels = percent_format(scale = 1)) + 
  scale_color_manual(values = c("0" = "#E42524", "1" = "#00ADA9"), labels = c("No Exemption", "With Exemption")) +
  labs(title = "Market Value Growth Rate (2016-2025)", 
       x = "Year", 
       y = "Growth Rate (%)", 
       color = "Exemption Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(yearly_market_value_filtered)

```

```{r}
# Calculate average growth rate by exemption status
growth_comparison <- yearly_market_value_filtered %>%
  group_by(exemption) %>%
  summarise(market_value_growth_rate = mean(market_value_growth_rate, na.rm = TRUE))

print(growth_comparison)

```


## 2.3 Summary Statistics for Each Parcel

Mean values for market_value, taxable_land, taxable_building, exempt_land, and exempt_building.
Growth rate (average annual growth).
Standard deviation (to measure volatility).

```{r}
# Summarize key metrics for each parcel: mean, growth rate, and standard deviation
residential_summary <- residential_assessment_combined %>%
  group_by(parcel_number) %>%
  summarise(
    avg_market_value = mean(market_value, na.rm = TRUE),
    avg_taxable_land = mean(taxable_land, na.rm = TRUE),
    avg_taxable_building = mean(taxable_building, na.rm = TRUE),
    avg_exempt_land = mean(exempt_land, na.rm = TRUE),
    avg_exempt_building = mean(exempt_building, na.rm = TRUE),
    growth_market_value = (last(market_value) - first(market_value)) / first(market_value) * 100,
    sd_market_value = sd(market_value, na.rm = TRUE),
    sd_taxable_land = sd(taxable_land, na.rm = TRUE),
    sd_taxable_building = sd(taxable_building, na.rm = TRUE)
  )

head(residential_summary)
```

```{r}
# Add exemption and residential status to the summary dataset
residential_summary <- residential_summary %>%
  left_join(
    residential_assessment_combined %>%
      select(parcel_number, exemption, is_residential, shape) %>%
      distinct(parcel_number, .keep_all = TRUE),  # Keep one row per parcel_number
    by = "parcel_number"
  )

head(residential_summary)
```

## 2.4 Analysis by Exemption Status
```{r}
# Summarize metrics by exemption status
residential_summary_analysis <- residential_summary %>%
  group_by(exemption) %>%
  summarise(
    avg_market_value = mean(avg_market_value, na.rm = TRUE),
    avg_taxable_land = mean(avg_taxable_land, na.rm = TRUE),
    avg_taxable_building = mean(avg_taxable_building, na.rm = TRUE),
    avg_exempt_land = mean(avg_exempt_land, na.rm = TRUE),
    avg_exempt_building = mean(avg_exempt_building, na.rm = TRUE),
    sd_market_value = mean(sd_market_value, na.rm = TRUE),
    sd_taxable_land = mean(sd_taxable_land, na.rm = TRUE),
    sd_taxable_building = mean(sd_taxable_building, na.rm = TRUE)
  ) 

# Convert the summary data to long format for visualization
library(tidyr)
summary_residential_long <- residential_summary_analysis %>%
  pivot_longer(
    cols = -exemption, 
    names_to = "metric", 
    values_to = "value"
  )
```



## 2.5 Visualizing Key Metrics by Exemption Status
```{r}
# Plot residential property metrics by exemption status
library(ggplot2)
custom_colors <- c("0" = "#E42524",  
                   "1" = "#00ADA9")  

ggplot(summary_residential_long, aes(x = metric, y = value, fill = as.factor(exemption))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors, labels = c("No Exemption", "With Exemption")) +
  labs(title = "Residential Property Metrics by Homestead Exemption Status (2015-2025)",
       x = "Metric", 
       y = "Value", 
       fill = "Exemption Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Plot Only "avg_market_value" and "sd_market_value"

```{r}
library(ggplot2)
library(dplyr)

summary_filtered <- summary_residential_long %>%
  filter(metric %in% c("avg_market_value", "sd_market_value"))

custom_colors <- c("0" = "#E42524",  
                   "1" = "#00ADA9") 

ggplot(summary_filtered, aes(x = metric, y = value, fill = as.factor(exemption), color = as.factor(exemption))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) + 
  geom_text(aes(label = round(value, 0)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 4.5, fontface = "bold") + 
  ylim(0, max(summary_filtered$value) * 1.2) +
  scale_fill_manual(values = custom_colors, labels = c("No Exemption", "With Exemption")) +
  scale_color_manual(values = custom_colors, guide = "none") +  
  labs(title = "Average and Standard Deviation of Market Value (2015-2025)",
       subtitle = "Comparison of Properties With and Without Exemption",
       x = "Metric", 
       y = "Value ($)", 
       fill = "Exemption Status") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 13, color = "gray40"))

```


# 3. Yearly Trends in Residential Data
## 3.1 Prepare Yearly Residential Data
```{r}
yearly_residential_data <- assessments2 %>%
  inner_join(cleaned_properties, by = "parcel_number") %>%
  filter(is_residential == 1)  # Keep only residential properties
```


## 3.2 Summarize Yearly Residential Data
```{r}
# Summarize yearly residential data by exemption status
yearly_residential_summary <- yearly_residential_data %>%
  group_by(year, exemption) %>%
  summarise(
    market_value = mean(market_value, na.rm = TRUE),
    taxable_land = mean(taxable_land, na.rm = TRUE),
    taxable_building = mean(taxable_building, na.rm = TRUE),
    exempt_land = mean(exempt_land, na.rm = TRUE),
    exempt_building = mean(exempt_building, na.rm = TRUE),
    .groups = "drop"
  )

```


## 3.3 Visualize Yearly Trends
```{r}
# Plot yearly changes in residential metrics
library(ggplot2)
library(tidyr)

# Convert the data to long format for better visualization
yearly_residential_long <- yearly_residential_summary %>%
  pivot_longer(
    cols = c(market_value, taxable_land, taxable_building, exempt_land, exempt_building),
    names_to = "metric",
    values_to = "value"
  )

# Filter data for years 2015 and later
yearly_residential_long_filtered <- yearly_residential_long %>%
  filter(year >= 2015)

custom_colors <- c("0" = "#E42524",  # Red (No Exemption)
                   "1" = "#00ADA9")  # Green (With Exemption)

# Generate individual plots for each metric
plots <- yearly_residential_long_filtered %>%
  split(.$metric) %>%
  lapply(function(df) {
    ggplot(df, aes(x = year, y = value, color = as.factor(exemption))) +
      geom_line(size = 1.2) +
      geom_point(size = 2) +
      scale_x_continuous(breaks = seq(2015, max(df$year), by = 1)) +
      scale_color_manual(values = custom_colors, labels = c("No Exemption", "With Exemption")) +
      labs(title = paste("Yearly Change in", unique(df$metric)),
           x = "Year", 
           y = "Value", 
           color = "Exemption Status") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  })


for (p in plots) {
  print(p)
}
```




# 1. Filter and Merge Residential Properties with Transfers Data
```{r}
transfers2 <- read.csv("~/Desktop/Penn/25spring/practicum/data/RTT_SUMMARY.csv")
colnames(transfers2)
table(transfers2$document_type)

```


```{r}
library(dplyr)

residential_properties <- cleaned_properties %>%
  filter(is_residential == 1)

residential_transfers <- residential_properties %>%
  left_join(transfers2, by = c("parcel_number" = "opa_account_num"))
```

# 2.Deed Type Analysis
```{r}
residential_transfers <- residential_transfers %>%
  mutate(document_type = ifelse(is.na(document_type), "Unknown", document_type))
```


```{r}
exemption_by_document_type <- residential_transfers %>%
  group_by(document_type) %>%
  summarise(
    total_count = n(),
    exemption_count = sum(exemption == 1),
    exemption_proportion = exemption_count / total_count * 100
  ) %>%
  arrange(desc(exemption_proportion))

print(exemption_by_document_type)
```


```{r}
## binomial model:DEED - DECEASED, SATISFACTION OF MORTGAGE, MORTGAGE, ASSIGNMENT OF MORTGAGE
document_exemption_model <- glm(exemption ~ document_type, data = residential_transfers, family = binomial)

summary(document_exemption_model)
```


# 3. Minimal Financial Consideration Analysis
```{r}
library(dplyr)

minimal_threshold <- 10 

minimal_financial_transfers <- residential_transfers %>%
  filter(total_consideration <= minimal_threshold)
```

```{r}

exemption_summary <- minimal_financial_transfers %>%
  group_by(exemption) %>%
  summarise(
    count = n(),
    proportion = count / nrow(minimal_financial_transfers) * 100
  )

print(exemption_summary)
```

```{r}
document_exemption_model <- glm(exemption ~ document_type, data = residential_transfers, family = binomial)

summary(document_exemption_model)
```

# 4. Recent(2y) Transfer Analysis
```{r}
library(dplyr)
library(lubridate)

# Convert recording_date to Date format
residential_transfers <- residential_transfers %>%
  mutate(recording_date = as.Date(recording_date, format="%Y-%m-%d"))

# Get the current year
current_year <- year(Sys.Date())
```


```{r}
# Filter and summarize recent transfers
residential_transfers_2y <- residential_transfers %>%
  filter(!is.na(recording_date)) %>%
  group_by(parcel_number) %>%
  summarise(
    latest_transfer_year = max(year(recording_date), na.rm = TRUE),  # Most recent transfer year
    exemption = first(exemption),  # Retain exemption status
    .groups = "drop"
  ) %>%
  mutate(
    latest_transfer_year = ifelse(is.infinite(latest_transfer_year), NA, latest_transfer_year)  # Handle infinite values
  )

# Mark properties with recent transfers (within 2 years)
residential_transfers_2y <- residential_transfers_2y %>%
  mutate(has_recent_transfer = ifelse(!is.na(latest_transfer_year) & latest_transfer_year >= (current_year - 2), 1, 0))

# View results
print(head(residential_transfers_2y))
```

```{r}
# Calculate recent transfer rates by exemption status
exemption_transfer_analysis_residential <- residential_transfers_2y %>%
  group_by(exemption) %>%
  summarise(
    avg_recent_transfer = mean(has_recent_transfer, na.rm = TRUE) * 100  # Convert to percentage
  )

# View results
print(exemption_transfer_analysis_residential)
```

On average, 20.07% of properties without a homestead exemption had a transfer in the past two years, compared to 16.17% of those with an exemption. This indicates that properties without exemptions tend to have a slightly higher likelihood of recent transactions.


```{r}
# Create a contingency table for exemption and recent transfers
exemption_transfer_table <- table(residential_transfers_2y$exemption, residential_transfers_2y$has_recent_transfer)

# Perform chi-square test
chi_test_result <- chisq.test(exemption_transfer_table)

# View test results
print(chi_test_result)
```

