---
title: "FP-spatial"
author: "Whitney Zhu"
date: "2025-03-31"
output: html_document
---

```{r}
library(readr)
library(dplyr)
fp_data_0505 <- read_csv("~/Downloads/fp_data_0505.csv")
glimpse(fp_data_0505)
```

```{r}
library(sf)

fp_sf_0505 <- st_as_sf(fp_data_0505, coords = c("lon_4326", "lat_4326"), crs = 4326)
```

```{r}
shapefile_path <- "~/Downloads/tl_2020_42101_tract20/tl_2020_42101_tract20.shp"

philly_shapefile <- st_read(shapefile_path)
```

```{r}
library(ggplot2)

ggplot() +
  geom_sf(data = philly_shapefile, fill = "white", color = "black") +  
  geom_sf(data = fp_sf_0505, color = "red", alpha = 0.5, size = 0.1) +       
  labs(
    title = "False Positive Properties in Philadelphia",
    subtitle = "Predicted as Exempted but Actually Not",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```
# Kernel Density Estimation
```{r}
ggplot() +
  geom_sf(data = philly_shapefile, fill = "white", color = "gray80") +
  stat_density2d(
    data = fp_data_0505, 
    aes(x = lon_4326, y = lat_4326, fill = ..level..), 
    geom = "polygon", alpha = 0.6, color = NA
  ) +
  scale_fill_viridis_c() +
  coord_sf() +
  labs(title = "Density of False Positive Properties in Philadelphia") +
  theme_minimal()

```

# DBSCAN
```{r}
fp_proj <- st_transform(fp_sf_0505, 26918)

coords <- st_coordinates(fp_proj)
```


```{r}
library(dbscan)

eps_val <- 300   
min_pts <- 20    

db_result <- dbscan(coords, eps = eps_val, minPts = min_pts)

fp_proj$cluster <- as.factor(db_result$cluster)
```

```{r}
ggplot(fp_proj) +
  geom_sf(aes(color = cluster), size = 0.2, alpha = 0.7) +
  scale_color_viridis_d(option = "plasma", name = "Cluster") +
  labs(
    title = "DBSCAN Clustering of False Positive Properties",
    subtitle = paste("ε =", eps_val, ", minPts =", min_pts),
    caption = "Cluster 0 = noise (not assigned to any cluster)"
  ) +
  theme_minimal()
```



01 KDE heatmap hotspots = Regions where the estimated probability density function is high.
It is smoother and continuous, showing density trends rather than clear cluster boundaries.
Even moderately dense but widespread areas can appear as "hot" in the visualization.

02 DBSCAN hotspots = Regions where at least minPts points are located within a distance of eps from each other.
It is stricter and tends to ignore areas that are high-density but more dispersed.
Only dense clusters of closely packed points are identified.


# spatial join to Tracts

```{r}

library(dplyr)
library(sf)

philly_shapefile <- st_transform(philly_shapefile, st_crs(fp_proj))
fp_with_tract <- st_join(fp_proj, philly_shapefile)

cluster_summary <- fp_with_tract %>%
  st_drop_geometry() %>%
  group_by(TRACTCE20, cluster) %>%
  summarise(count = n(), .groups = "drop")

print(cluster_summary)
```

```{r}
tract_fp_stats <- cluster_summary %>%
  group_by(TRACTCE20) %>%
  summarise(
    total_fp = sum(count),
    n_clusters = n_distinct(cluster)
  ) %>%
  arrange(desc(total_fp))

head(tract_fp_stats)
```

```{r}
philly_tracts_joined <- left_join(philly_shapefile, tract_fp_stats, by = "TRACTCE20")
```

```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(scales)

philly_tracts_joined <- left_join(philly_shapefile, tract_fp_stats, by = "TRACTCE20")

top10_tracts <- philly_tracts_joined %>%
  arrange(desc(total_fp)) %>%
  slice_head(n = 10) %>%
  pull(TRACTCE20)

philly_tracts_joined$highlight_top10 <- philly_tracts_joined$TRACTCE20 %in% top10_tracts

ggplot() +
  geom_sf(data = philly_tracts_joined, fill = "white", color = "gray70", size = 0.3) +
  geom_sf(data = filter(philly_tracts_joined, !is.na(total_fp)),
          aes(fill = total_fp), color = NA) +
  
  scale_fill_gradient(low = "#00ADA9", high = "#E42524", name = "False Positive Count") +
  geom_sf(data = filter(philly_tracts_joined, highlight_top10),
          fill = NA, color = "#E42524", linewidth = 0.6) +
  labs(
    title = "Missed Exemption Opportunities by Census Tract",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )
```

```{r}
library(dplyr)

top10_tracts <- tract_fp_stats %>%
  arrange(desc(total_fp)) %>%
  slice(1:10)

print(top10_tracts)

library(knitr)
kable(top10_tracts, caption = "Top 10 Census Tracts by Predicted but Unclaimed Exemptions")

```





# Moran's I
```{r}
install.packages("spdep")  
library(spdep)
```

```{r}
philly_moran <- philly_tracts_joined %>% 
  filter(!is.na(total_fp))

tract_nb <- poly2nb(philly_moran)
tract_weights <- nb2listw(tract_nb, style = "W", zero.policy = TRUE)
```

```{r}
local_moran <- localmoran(
  x = philly_moran$total_fp,
  listw = tract_weights,
  zero.policy = TRUE
)

philly_moran$Ii <- local_moran[, 1]      # Local Moran's I
philly_moran$Z_Ii <- local_moran[, 4]    # 标准化 I
philly_moran$p_value <- local_moran[, 5] # p-value

```

```{r}
fp_mean <- mean(philly_moran$total_fp)

philly_moran <- philly_moran %>%
  mutate(
    lag_fp = lag.listw(tract_weights, total_fp),
    cluster_type = case_when(
      total_fp > fp_mean & lag_fp > fp_mean & p_value < 0.05 ~ "High-High",
      total_fp < fp_mean & lag_fp < fp_mean & p_value < 0.05 ~ "Low-Low",
      total_fp > fp_mean & lag_fp < fp_mean & p_value < 0.05 ~ "High-Low",
      total_fp < fp_mean & lag_fp > fp_mean & p_value < 0.05 ~ "Low-High",
      TRUE ~ "Not Significant"
    )
  )
```

```{r}
library(dplyr)

philly_moran %>%
  count(cluster_type)
```

```{r}
# 计算 High-High tracts 中总的 false positive 数量
philly_moran %>%
  filter(cluster_type == "High-High") %>%
  summarise(total_fp_in_highhigh = sum(total_fp, na.rm = TRUE))

```



```{r}
library(ggplot2)

ggplot(philly_moran) +
  geom_sf(aes(fill = cluster_type), color = "gray90", size = 0.2) +
  scale_fill_manual(
    values = c(
      "High-High" = "#E42524",
      "Low-Low" = "#00ADA9",
      "High-Low" = "#F39C12",
      "Low-High" = "#9B59B6",
      "Not Significant" = "lightgray"
    ),
    name = "LISA Cluster"
  ) +
  labs(
    title = "LISA Cluster Map (Local Moran's I)",
    subtitle = "Clusters of Missed Exemption Opportunities"
  ) +
  theme_minimal()

```

```{r}
library(ggplot2)

ggplot(philly_moran) +
  geom_sf(aes(fill = cluster_type), color = "gray90", size = 0.2) +
  scale_fill_manual(
    values = c(
      "High-High" = "#E42524"
    ),
    name = "LISA Cluster"
  ) +
  labs(
    title = "LISA Cluster Map (Local Moran's I)",
    subtitle = "Clusters of Missed Exemption Opportunities"
  ) +
  theme_minimal()

```

```{r}
highhigh_tracts <- philly_moran %>%
  filter(cluster_type == "High-High")

highhigh_union <- highhigh_tracts %>%
  st_union() %>%         
  st_sf()               

st_write(highhigh_union, "highhigh_union_0505.geojson", driver = "GeoJSON")
```







```{r}
library(sf)
library(dplyr)

highhigh_sf <- philly_moran %>% 
  filter(cluster_type == "High-High")

highhigh_sf <- highhigh_sf %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(lat = st_coordinates(centroid)[, 2])

lat_mid <- median(highhigh_sf$lat, na.rm = TRUE)

highhigh_summary <- highhigh_sf %>%
  mutate(region = ifelse(lat >= lat_mid, "North Group", "South Group")) %>%
  group_by(region) %>%
  summarise(
    num_tracts = n(),
    total_fp = sum(total_fp, na.rm = TRUE)
  )

print(highhigh_summary)
```




```{r}
ggplot(philly_moran) +
  geom_sf(aes(fill = p_value), color = "white") +
  scale_fill_viridis_c(option = "C", direction = -1) +
  labs(title = "P-Values from Local Moran's I", fill = "P-Value") +
  theme_minimal()
```

