---
title: "Optimizing Entitlement Assistance Outreach to Homeowners"
author: "Claudia Low, Haoyu Zhu, Rachel Midgett, Wenjun Zhu"
date: "2025 Spring"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)

library(tidycensus)
library(tidyverse)
library(dplyr)
library(sf)
library(spdep)
library(caret)
library(ckanr)
library(FNN)
library(grid)
library(gridExtra)
library(ggcorrplot) 
library(corrr)    
library(kableExtra)
library(jtools)   
library(ggstance)
library(ggpubr) 
library(broom.mixed)
library(RColorBrewer)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(corrr)
library(classInt)
library(stargazer)
library(RSocrata)
library(kableExtra)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(lubridate)
library(stargazer)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(scales)
library(car)
library(data.table)
library(leaflet)


options(scipen = 999)
root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

# Inflation adjust
inflationAdjust2017 <- 1.0213
inflationAdjust2018 <- 1.0244

# Set up theme and options
colors <- c("0" = "#e42524", "1" = "#00ADA9")

map_theme <- theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

options(tigris_use_cache = TRUE)
options(scipen = 999)
```

# Introduction

This report provides a detailed workflow of the project on Homestead Tax Exemption entitlement assisstance outreach, for the City of Philadelphia Office of Philly Stat 360 and Office of Information Technology. The aim of the project is design an algorithm-driven outreach campaign that can cost effectively identify homeowners who are likely to be eligible for the [Homestead Tax Exemption](https://www.phila.gov/services/payments-assistance-taxes/taxes/property-and-real-estate-taxes/get-real-estate-tax-relief/get-the-homestead-exemption/) but are not participating in the program. The project aims to allow our clients to understand where these properties are located, potential outreach strategies, and the associated costs and benefits.

These relevant properties who are identified as most likely eligible for the Homestead Exemption but not taking up the program,  are also thought to be more likely to be subject to “tangled titles,” or family-rental arrangements that require an affidavit to waive need for a rental license.

## Background on Property Tax in Philadelphia
Property tax in Philadelphia is 1.3998% of the property value, as assessed by the Office of Property Assessment,for the 2025 taxx year. This is made up of 0.6159% (City of Philadelphia) and 0.7839% (School District)
The taxes are due March 31st yearly.

## Background on Homestead Exemption
The Homestead Exemption reduces the taxable portion of a homeowner's property assessment by up to $100,000, saving up to $1,399 on real estate taxes annually. The bill signed aimed to lessen the financial burden of new property assessments on Philadelphia homeowners, whose property values increased by an average of 31% after the city delayed the annual calculations for three years due to the pandemic.
Eligibility for the Homestead Exemption is as follows:
•	you must own the property and use it as your primary residence
•	no age or income restrictions
•	Not used exclusively for business purposes or as rental units (a percentage is fine)

A homeowner is Ineligible if a homeowner is already enrolled in these alternative real estate tax relief/abatement programs:
•	Longtime Owner Occupants Program (LOOP), an	income-based program for homeowners who experience a substantial increase in their property assessment.
•	10-year residential tax abatement program, although one can only apply for Homestead Exemption after the abatement is over

Programs that can be used in conjunction with the homestead exemption include 
•	Owner-Occupied Real Estate Tax Payment Agreement (OOPA)
•	Senior Citizen Real Estate Tax Freeze
•	Low-Income Real Estate Tax Freeze
•	Real Estate Tax Installment Plan
• Tax Credits for Active-Duty Reserve and National Guard Members

## Tangled Titles

## Significance of Outreach 

# Data
## Dataset Overview
The primary dataset used is the Property and Assessment History publicly available for download on OpenDataPhilly. 
Six relevant datasets are merged with this primary dataset with common identifying keys such as the parcel number in order to include useful predictor variabels in the model predicting for homeowners most likely eligible but not currently enrolled in the Homestead Exemption.

## Exploratory Analysis
### Property and Assessment History
Every observation in the Property and Assessment History dataset is one property in Philadelphia, with a total of 584,049 properties and 79 features. As this dataset is updated daily, the one used for this project is updated as of 31 January 2025. 

#### Creation of dependent variable
There is a column 'homestead_exemption' within this dataset which indicates the taxable portion amount removed from the property assessment of the house. It should be noted that there are 14 properties that had a homestead exemption larger then $100,000, the maximum possible amount, which is suspected to be a clerical error and has been flagged to the PhillyStat360 team. The dependent variable for the model is derived from this feature by creating a binary variable on whether or not the property is currently enrolled in the homestead exemption program, indicated by a non-zero value. There are 246,853 properties with a homestead exemption.
```{r}
properties <- fread("Data/opa_properties_public.csv")
filtered_properties <- properties %>%
  mutate(exemption = ifelse(homestead_exemption == 0, 0, 1))
```

#### Filtering to residential properties
```{r}
filtered_properties <- filtered_properties %>% mutate(is_residential = ifelse(zoning %in% c(
  "RM1", "RM2", "RM3", "RM4",
  "RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RSA6", 
  "RSD1", "RSD2", "RSD3", 
  "RM1|RSA5", "RSD1|RSD3", "RSA5|RSA5",
  "RTA1", 
  "CMX1", "CMX2", "CMX2.5", "CMX3", "CMX4", "CMX5", "IRMX"), 1, 0))

# Look into exemption status by zoning code
exemptionbyzoning <- filtered_properties %>%
  group_by(zoning, exemption) %>%
  summarise(count = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = exemption, values_from = count, values_fill = list(count = 0)) %>%
  rename(No_Exemption = `0`, Exemption = `1`)

# Look into number of blank / NA zoning codes
num_blank_zoning <- properties %>%
  filter(zoning == "" | is.na(zoning) | str_trim(zoning) == "") %>%
  nrow()

print(num_blank_zoning)

residential_properties <- filtered_properties %>% filter(is_residential == 1)

```
#### transform to geodata

```{r}
properties_sf <- st_as_sf(residential_properties, wkt = "shape", crs = 2272)
```

#### Homestead Rate by Census Tract
```{r}
census_tracts <- st_read("Data/phila_census1.gpkg")
```

```{r}
census_tracts <- st_transform(census_tracts, 2272)
properties_tract <- st_join(properties_sf, census_tracts)

# Create tract summary
tract_summary <- properties_tract %>%
  group_by(GEOID) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100,
    .groups = "drop"
  ) %>%
  st_drop_geometry()

# Create final enriched dataset
census_tracts_enriched <- census_tracts %>%
  left_join(tract_summary, by = "GEOID")
```


```{r}
# Transform data to WGS84 (required for leaflet)
census_tracts_wgs84 <- st_transform(census_tracts_enriched, 4326)

# Create interactive map
leaflet(census_tracts_wgs84) %>%
  addTiles() %>%  # Add OpenStreetMap base map
  addPolygons(
    fillColor = ~colorNumeric(
      palette = "viridis",
      domain = c(0, 100)
    )(pct_homestead),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    popup = ~paste(
      "Census Tract:", GEOID, "<br>",
      "Homestead %:", round(pct_homestead, 1), "<br>",
      "Population Density:", round(pop_density, 0), "<br>",
      "Total Properties:", total_properties, "<br>",
      "Median Income:", scales::dollar(median_income)
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("viridis", domain = c(0, 100)),
    values = ~pct_homestead,
    title = "% Homestead Exemption",
    opacity = 0.7
  )

```
```{r}
census_tracts_filtered <- census_tracts_enriched %>%
  filter(pop_density > 0 & total_properties >= 30)  # 30 properties minimum

census_tracts_invalid <- census_tracts_enriched %>%
  filter(pop_density == 0 | total_properties < 30)  # Include low property counts in "invalid"


# Transform both to WGS84
census_tracts_invalid_wgs84 <- st_transform(census_tracts_invalid, 4326)
census_tracts_filtered_wgs84 <- st_transform(census_tracts_filtered, 4326)

# Create map with both layers
leaflet() %>%
  addTiles() %>%
  # Add invalid tracts first (in gray)
  addPolygons(data = census_tracts_invalid_wgs84,
    fillColor = "gray",
    fillOpacity = 0.5,
    weight = 1,
    color = "white",
    popup = "No data available"
  ) %>%
  # Add valid tracts with your original styling
  addPolygons(data = census_tracts_filtered_wgs84,
    fillColor = ~colorNumeric(
      palette = "viridis",
      domain = c(0, 85)
    )(pct_homestead),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    popup = ~paste(
    "Census Tract:", GEOID, "<br>",
    "Homestead %:", round(pct_homestead, 1), "%<br>",
    "Total Properties:", total_properties,
    ifelse(total_properties < 100, 
           "<br><i style='color:red'>Note: Low property count may affect reliability</i>", 
           "")
  )

  )%>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("viridis", domain = c(0, 85)),
    values = census_tracts_filtered_wgs84$pct_homestead,
    title = "% Homestead Exemption",
    opacity = 0.7,
    labFormat = labelFormat(suffix = "%")
  ) %>%
  # Add legend for gray areas
  addLegend(
    position = "bottomright",
    colors = "gray",
    labels = "No Data Available",
    opacity = 0.5
  )


```


```{r}
# Create histogram of homestead exemption distribution
census_hist <- ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), # Filter out zero population tracts
       aes(x = pct_homestead)) +
  geom_histogram(
    binwidth = 5,
    fill = "#008d8a",
    color = "white"
  ) +
  labs(
    title = "Distribution of Homestead Exemption Rates Across Philadelphia Census Tracts",
    subtitle = "Excluding Zero Population Density Tracts",
    x = "Percentage of Properties with Homestead Exemption",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14)
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10))

census_hist
```
The distribution shows a roughly normal shape with most tracts clustered between 30-50%
There's a notable drop-off below 30% in the number of tracts
The histogram shows relatively few tracts with rates below 20%

Therefore, census tracts with homestead exemption rates below 30% could be considered to have low enrollment and might warrant targeted outreach or investigation into barriers to participation, assuming they are primarily residential areas and not institutional/special use tracts.

There's a notable drop-off below 30% in the number of tracts. Therefore, census tracts with homestead exemption rates below 30% could be considered to have low enrollment and might warrant targeted outreach or investigation into barriers to participation, assuming they are primarily residential areas.

```{r}
homestead_pattern <- ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = cut(pct_homestead, 
              breaks = c(0, 20, 30, 40, 50, 60, 100),
              labels = c("<20%", "20-30%", "30-40%", "40-50%", "50-60%", ">60%")))) +
  scale_fill_viridis_d(
    name = "Homestead\nExemption Rate",
    na.value = "gray80",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    title = "Homestead Exemption Rates Across Philadelphia",
    subtitle = "By Census Tract (Excluding Zero Population Areas)",
    caption = "Gray areas indicate zero population density tracts"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_blank(),
    legend.text = element_text(size = 14)
  )
```


```{r}
homestead_pattern
#ggsave("outputs/homestead-exemption-pattern.png", homestead_pattern, width = 10, height = 6)
```

```{r}
# Basic summary statistics
summary(census_tracts_enriched$pop_density)

# More detailed statistics
census_tracts_enriched %>%
  summarise(
    mean_density = mean(pop_density, na.rm = TRUE),
    median_density = median(pop_density, na.rm = TRUE),
    q1 = quantile(pop_density, 0.25, na.rm = TRUE),
    q3 = quantile(pop_density, 0.75, na.rm = TRUE)
  )

# Visual distribution
ggplot(census_tracts_enriched, aes(x = pop_density)) +
  geom_histogram(binwidth = 1000) +
  theme_minimal() +
  labs(title = "Distribution of Population Density in Philadelphia Census Tracts",
       x = "Population Density (per square mile)",
       y = "Count of Census Tracts")

```


```{r}
ggplot(census_tracts_enriched, aes(x = pop_density)) +
  geom_histogram(binwidth = 1000) +
  theme_minimal() +
  labs(title = "Distribution of Population Density in Philadelphia Census Tracts",
       x = "Population Density (per square mile)",
       y = "Count of Census Tracts")
```


```{r}
homestead_pattern <- ggplot(census_tracts_enriched %>% 
                           mutate(pct_homestead = case_when(
                             pop_density < 2000 | total_properties < 100 ~ NA_real_,
                             TRUE ~ pct_homestead))) +
  geom_sf(aes(fill = cut(pct_homestead, 
              breaks = c(0, 20, 30, 40, 50, 60, 100),
              labels = c("<20%", "20-30%", "30-40%", "40-50%", "50-60%", ">60%")))) +
  scale_fill_viridis_d(
    name = "Homestead\nExemption Rate",
    na.value = "gray80",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    title = "Homestead Exemption Rates Across Philadelphia",
    subtitle = "By Census Tract",
    caption = "Gray tracts indicate population density < 2,000 per sq. mile or fewer than 100 properties"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_blank(),
    legend.text = element_text(size = 14)
  )

```

```{r}
homestead_pattern
```

```{r}
# Distribution of pct_homestead
summary(tract_summary$pct_homestead)

# Map for low rates (under 30%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density > 0 & pct_homestead < 30, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Under 30%)",
    na.value = "gray80",
    limits = c(0, 30),
    breaks = seq(0, 30, by = 5)
  ) +
  labs(
    title = "Low Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Below 30% Enrollment",
    caption = "Gray areas: Zero population density or rates ≥ 30%"
  ) +
  map_theme

# Table for low rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density > 0 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

# Map for high rates (over 60%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density > 0 & pct_homestead > 60, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Over 60%)",
    na.value = "gray80",
    limits = c(60, 82),
    breaks = seq(60, 80, by = 5)
  ) +
  labs(
    title = "High Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Above 60% Enrollment",
    caption = "Gray areas: Zero population density or rates ≤ 60%"
  ) +
  map_theme

# Table for high rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density > 0 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

```

```{r}
# Distribution of pct_homestead with new criteria
summary(tract_summary$pct_homestead[tract_summary$pop_density >= 2000 & tract_summary$total_properties >= 100])

# Map for low rates (under 30%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density >= 2000 & total_properties >= 100 & pct_homestead < 30, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Under 30%)",
    na.value = "gray80",
    limits = c(0, 30),
    breaks = seq(0, 30, by = 5)
  ) +
  labs(
    title = "Low Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Below 30% Enrollment",
    caption = "Gray areas: Low density (<2000/sq mi), low property count (<100), or rates ≥ 30%"
  ) +
  map_theme

# Table for low rates
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & total_properties >= 100 & pct_homestead < 30) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(pct_homestead)

low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%)"
  )

# Map for high rates (over 60%)
ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = ifelse(pop_density >= 2000 & total_properties >= 100 & pct_homestead > 60, 
              pct_homestead, NA))) +
  scale_fill_viridis_c(
    name = "% Homestead\nExemption\n(Over 60%)",
    na.value = "gray80",
    limits = c(60, 82),
    breaks = seq(60, 80, by = 5)
  ) +
  labs(
    title = "High Homestead Exemption Rates in Philadelphia",
    subtitle = "Census Tracts Above 60% Enrollment",
    caption = "Gray areas: Low density (<2000/sq mi), low property count (<100), or rates ≤ 60%"
  ) +
  map_theme

# Table for high rates
high_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & total_properties >= 100 & pct_homestead > 60) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(desc(pct_homestead))

high_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(desc(pct_homestead)) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with High Homestead Exemption Rates (>60%)"
  )

```

```{r}
census_tracts_enriched <- census_tracts_enriched %>%
  mutate(
    owner_occ_rate = (census_tracts_enriched$owner_hh / census_tracts_enriched$occupied_units) * 100
  )
    
    
low_enrollment_tracts <- census_tracts_enriched %>%
  filter(pop_density >= 2000 & 
         total_properties >= 100 & 
         pct_homestead < 30 &
         owner_occ_rate > 40) %>%
  select(GEOID, pct_homestead, pop_density, total_properties) %>%
  arrange(pct_homestead)

# For the map visualization
low_enrollment_tracts_map <- ggplot(census_tracts_enriched) +
  geom_sf(aes(fill = case_when(
    pop_density >= 2000 & 
    total_properties >= 100 & 
    pct_homestead < 30 &
    owner_occ_rate > 40 ~ "#f4aa9e",
    TRUE ~ "gray80"
  ))) +
  scale_fill_identity() +
  labs(
    title = "Low Homestead Exemption Tracts",
    subtitle = "Tracts with <30% Homestead Rate & >40% Owner Occupancy",
    caption = "Gray areas do not meet filtering criteria"
  ) +
  map_theme




low_enrollment_tracts_map
#ggsave("outputs/homestead-exemption-low_enrollment_tracts.png", low_enrollment_tracts_map, width = 10, height = 6)
```
```{r}
# For the table
low_enrollment_tracts %>%
  st_drop_geometry() %>% 
  arrange(pct_homestead) %>%
  kable(
    col.names = c("Census Tract", "% Homestead", "Population Density", "Total Properties"),
    digits = 2,
    caption = "Census Tracts with Low Homestead Exemption Rates (<30%) and High Owner Occupancy (>40%)"
  )
```

```{r}
ggplot(census_tracts_enriched %>% 
       filter(pop_density >= 2000), # Using your established density threshold
       aes(x = owner_occ_rate)) +
  geom_histogram(
    binwidth = 5,
    fill = "#e42524",  # Using your established red color
    alpha = 0.8,
    color = "white"
  ) +
  labs(
    title = "Distribution of Owner Occupancy Rates Across Philadelphia Census Tracts",
    subtitle = "Excluding Low Density Areas (<2,000 per square mile)",
    x = "Owner Occupancy Rate (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10))

```


#### Predictor Variables 1 & 2: Eligibility Characteristics
The first predictor variable is based off the key eligibility criteria of the Homestead Exemption program -- whether a homeowner resides long-term in the property itself. The closest proxy of this is whether the mailing street ('mailing_street') address is the same as the property street address ('location'), as the mailing address of the owner is often where the owner lives long-term.

The second predictor variable is based off the ineligibility for the homestead exemption if the homeowner is already enrolled in particular exisitng tax programs such as the LOOP and the residential tax abatement. This is determined if there is an existing tax relief that exempts a portion of the building from tax. If there is a non-zero value for 'exempt-building' and the property is not currently enrolled in the homestead exemption, it is potentially already enrolled in LOOP or in residential tax abatement.

```{r}
residential_properties <- residential_properties %>% mutate(same_address = ifelse(mailing_street == location, 1, 0))

ggplot(residential_properties, aes(x = same_address, fill = as.factor(exemption))) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = colors, labels = c("No Exemption", "With Exemption")) +
    labs(title = "Mailing Address Matches Property Address", x = "Address Match", y = "Count", fill = "Homestead Exemption") +
    theme_minimal(base_size = 14) +
    theme(panel.grid.major = element_line(color = "grey90"),
          panel.grid.minor = element_blank(),
          legend.position = "bottom",
         plot.title = element_text(vjust = 0.5))

residential_properties <- residential_properties %>% mutate(potential_otherprog = ifelse(exempt_building > 0 & exemption == 0, 1, 0))

```

#### Predictor Variables 3 & 4: Property Characteristics
Depth of the property as well as mean total property area 
```{r}
residential_properties <- residential_properties %>% mutate(is_deep = ifelse(depth > 150, 1, 0)) 

ggplot(residential_properties %>% 
         filter(!is.na(is_deep)), 
       aes(x = factor(is_deep), fill = factor(exemption))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = colors, labels = c("No Exemption", "With Exemption")) +
  labs(title = "Property Depth by Homestead Exemption", 
       x = "Depth > 300", 
       y = "Count", 
       fill = "Exemption") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major = element_line(color = "grey90"),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(vjust = 0.5))
```



#### Predictor Variables 5 & 6: Petentail Commercial Activity

```{r}
business_license<-st_read("Data/business_licenses.geojson")
```

```{r}
rental_license<-business_license%>%
  filter(licensetype=="Rental",
         #rentalcategory=="Residential Dwellings",
         licensestatus %in% c("Active"))%>%
  select(opa_account_num)%>%
  st_drop_geometry()%>%
  distinct() %>% 
  filter(!is.na(opa_account_num)) %>% 
  mutate(rental_license = 1)  
```

```{r}
properties_rental <- properties_sf %>%
  mutate(parcel_number = as.character(parcel_number))%>%
  left_join(rental_license, by = c("parcel_number" = "opa_account_num"))%>%
  mutate(rental_license = replace_na(rental_license, 0))
```

```{r}
ggplot(properties_rental, aes(x = factor(rental_license), fill = factor(exemption))) +
  geom_bar(position = "dodge") +  
  geom_text(stat = "count", aes(label = ..count.., color = factor(exemption)),  
            position = position_dodge(width = 0.9),  
            vjust = -0.5,  
            size = 3) +  
  scale_fill_manual(values = c("0" = "#e42524", "1" = "#00ADA9"),  
                    labels = c("0" = "No Exemption", "1" = "With Exemption")) +  
  scale_color_manual(values = c("0" = "#e42524", "1" = "#00ADA9")) +  
  labs(title = "Rental Licenses Metrics by Homestead Exemption Status",
       x = "Rental Licenses",
       y = "Count",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(legend.position = "none") 
```

```{r}
commercial_license<-business_license%>%
  filter(licensestatus %in% c("Active"))%>%
  filter(licensetype %in% c(
    "Food Caterer",
    "Food Establishment, Retail Perm Location (Large)",
    "Food Establishment, Retail Permanent Location",
    "Food Manufacturer / Wholesaler",
    "Food Preparing and Serving",
    "Food Preparing and Serving (30+ SEATS)",
    "Motor Vehicle Repair / Retail Mobile Dispensing",
    "Pawn Shop",
    "Precious Metal Dealer",
    "Public Garage / Parking Lot",
    "Residential Property Wholesaler",
    "Tire Dealer",
    "Tow Company",
    "Vacant Commercial Property"
  ))%>%
  select(opa_account_num)%>%
  st_drop_geometry()%>%
  distinct() %>% 
  filter(!is.na(opa_account_num)) %>% 
  mutate(commercial_license = 1) 
```

```{r}
properties_commercial <- properties_sf %>%
  mutate(parcel_number = as.character(parcel_number))%>%
  left_join(commercial_license, by = c("parcel_number" = "opa_account_num"))%>%
  mutate(commercial_license = replace_na(commercial_license, 0))
```

```{r}
ggplot(properties_commercial, aes(x = factor(commercial_license), fill = factor(exemption))) +
  geom_bar(position = "dodge") +  
  geom_text(stat = "count", aes(label = ..count.., color = factor(exemption)),  
            position = position_dodge(width = 0.9),  
            vjust = -0.5,  
            size = 3) +  
  scale_fill_manual(values = c("0" = "#e42524", "1" = "#00ADA9"),  
                    labels = c("0" = "No Exemption", "1" = "With Exemption")) +  
  scale_color_manual(values = c("0" = "#e42524", "1" = "#00ADA9")) +  
  labs(title = "Commercial Licenses Metrics by Homestead Exemption Status",
       x = "Commercial Licenses (Exclude Rental)",
       y = "Count",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(legend.position = "none") 
```

#### Predictor Variables 7 & 8: Tax Balance


```{r}
balances <- read.csv("Data/real_estate_tax_balances_census_tract.csv")
```

```{r}
balance_sf <- census_tracts_enriched %>% 
  left_join(balances %>% select(census_tract,balance,num_props), 
            by = c("GEOID" = "census_tract")) 
```

```{r}
ggplot(balance_sf) +
  geom_sf(aes(fill = balance), color = "white", size = 0.1) + 
  scale_fill_gradientn(colors = c("#00ADA9","#e3f9f7","#f4aa9e", "#e42524"),
                       limits = range(balance_sf$balance, na.rm = TRUE),
                       breaks = range(balance_sf$balance, na.rm = TRUE)
                       ) + 
  labs(title = "Total Tax Balance by Census Tract",
       fill = "Price") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 6),
    axis.title = element_text(size = 6),
    axis.text = element_blank(),
    legend.text = element_text(size = 6),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```

```{r}
# Visualize the relationship
ggplot(balance_sf, aes(x = balance, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(
    title = "Principle vs. Homestead Participation",
    x = "Total Principle ($)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()
```

```{r}
#distinct
balance_distinct <- balances %>%
  mutate(census = as.numeric(substr(census_tract, 6, 9)))%>%
  st_drop_geometry()%>%
  group_by(census) %>% 
  summarise(avg_balance=sum(balance,na.rm=TRUE)/sum(num_props, na.rm = TRUE),
            sum_balance=sum(balance,na.rm=TRUE),
            tax_props=sum(num_props,na.rm=TRUE)) %>%
  ungroup()
```

```{r}
properties_number<-properties_sf%>%
  select(census_tract,exemption,parcel_number)%>%
  mutate(prop=1)%>%
  st_drop_geometry()%>%
  group_by(census_tract)%>%
  summarise(total_props=sum(prop))%>%
  ungroup()

properties_sf_number<-properties_sf%>%
  left_join(properties_number,by="census_tract")

```


```{r}
properties_balance <- properties_sf_number %>%
  select(objectid, census_tract,exemption,parcel_number,total_props)%>%
  left_join(balance_distinct, by = c("census_tract" = "census"))%>%
  mutate(avg_balance = replace(avg_balance, is.na(avg_balance), 0),
         sum_balance = replace(sum_balance, is.na(sum_balance), 0),
         tax_props=replace(tax_props,is.na(tax_props),0))%>%
  mutate(tax_rate=tax_props/total_props)
  
```

```{r}
avg_values <- properties_balance %>%
  st_drop_geometry()%>%
  group_by(exemption) %>%
  summarise(
    'Total Tax Balance (In 100,000)' = mean(sum_balance, na.rm = TRUE)/100000,
    #Tax_Props=mean(tax_props,na.rm=TRUE),
    '% of Properties with Tax Balance'=mean(tax_rate,na.rm=TRUE)*100
  ) %>%
  pivot_longer(cols = -exemption, names_to = "variable", values_to = "mean_value")

ggplot(avg_values, aes(x = variable, y = mean_value, fill = as.factor(exemption))) +
  geom_bar(stat = "identity", width = 0.5, position = position_dodge(width = 0.6)) +

  geom_text(aes(label = round(mean_value, 4), color = as.factor(exemption)), 
            position = position_dodge(width = 0.6), 
            vjust = -0.5, size = 5) +

  labs(title = "Mean of Tax Variables by Exemption Status",
       x = "Metrics",
       y = "Mean Value",
       fill = "Exemption Status") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  
  scale_fill_manual(values = c("#e42524", "#00ADA9"), labels = c("No Exemption", "With Exemption")) +
  scale_color_manual(values = c("#e42524", "#00ADA9"), guide = "none") 
```

#### Predictor Variables: Assessment History
```{r}
# Load and inspect the assessments dataset
assessments2 <- read.csv("Data/assessments.csv")
```

```{r}
colnames(assessments2)
head(assessments2, 10)

# Merge properties and assessments data by parcel_number
cleaned_properties <- filtered_properties %>%
  select(parcel_number, exemption, is_residential, shape)
assessment_combined <- assessments2 %>%
  left_join(cleaned_properties, by = "parcel_number")

head(assessment_combined)
```

##### Residential Property History Analysis

```{r}
# Filter Residential Data
# Filter the combined dataset to include only residential properties
residential_assessment_combined <- assessment_combined %>%
  filter(is_residential == 1)
head(residential_assessment_combined)
colnames(residential_assessment_combined)
```

```{r}
# Market Value Growth Rate by Exemption Status (2016-2025)

# Calculate yearly market value and growth rate by exemption status
yearly_market_value <- residential_assessment_combined %>%
  group_by(year, exemption) %>%
  summarise(total_market_value = sum(market_value, na.rm = TRUE), .groups = 'drop')


yearly_market_value <- yearly_market_value %>%
  arrange(exemption, year) %>%
  group_by(exemption) %>%
  mutate(market_value_growth_rate = (total_market_value - lag(total_market_value)) / lag(total_market_value) * 100)

# Filter data for the years 2016-2025
yearly_market_value_filtered <- yearly_market_value %>%
  filter(year >= 2016 & year <= 2025)

# Plot the market value growth rate
ggplot(yearly_market_value_filtered, aes(x = year, y = market_value_growth_rate, color = as.factor(exemption))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, by = 1)) + 
  scale_y_continuous(labels = percent_format(scale = 1)) + 
  scale_color_manual(values = c("0" = "#E42524", "1" = "#00ADA9"), labels = c("No Exemption", "With Exemption")) +
  labs(title = "Market Value Growth Rate (2016-2025)", 
       x = "Year", 
       y = "Growth Rate (%)", 
       color = "Exemption Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(yearly_market_value_filtered)
```

```{r}
# Calculate average growth rate by exemption status
growth_comparison <- yearly_market_value_filtered %>%
  group_by(exemption) %>%
  summarise(market_value_growth_rate = mean(market_value_growth_rate, na.rm = TRUE))

print(growth_comparison)

```

Mean values for market_value, taxable_land, taxable_building, exempt_land, and exempt_building.
Growth rate (average annual growth).
Standard deviation (to measure volatility).

```{r}
# Summarize key metrics for each parcel: mean, growth rate, and standard deviation
residential_summary <- residential_assessment_combined %>%
  group_by(parcel_number) %>%
  summarise(
    avg_market_value = mean(market_value, na.rm = TRUE),
    avg_taxable_land = mean(taxable_land, na.rm = TRUE),
    avg_taxable_building = mean(taxable_building, na.rm = TRUE),
    avg_exempt_land = mean(exempt_land, na.rm = TRUE),
    avg_exempt_building = mean(exempt_building, na.rm = TRUE),
    growth_market_value = (last(market_value) - first(market_value)) / first(market_value) * 100,
    sd_market_value = sd(market_value, na.rm = TRUE),
    sd_taxable_land = sd(taxable_land, na.rm = TRUE),
    sd_taxable_building = sd(taxable_building, na.rm = TRUE)
  )
```

```{r}
# Add exemption and residential status to the summary dataset
residential_summary <- residential_summary %>%
  left_join(
    residential_assessment_combined %>%
      select(parcel_number, exemption, is_residential, shape) %>%
      distinct(parcel_number, .keep_all = TRUE),  # Keep one row per parcel_number
    by = "parcel_number"
  )

head(residential_summary)
```

```{r}
# Analysis by Exemption Status
# Summarize metrics by exemption status
residential_summary_analysis <- residential_summary %>%
  group_by(exemption) %>%
  summarise(
    avg_market_value = mean(avg_market_value, na.rm = TRUE),
    avg_taxable_land = mean(avg_taxable_land, na.rm = TRUE),
    avg_taxable_building = mean(avg_taxable_building, na.rm = TRUE),
    avg_exempt_land = mean(avg_exempt_land, na.rm = TRUE),
    avg_exempt_building = mean(avg_exempt_building, na.rm = TRUE),
    sd_market_value = mean(sd_market_value, na.rm = TRUE),
    sd_taxable_land = mean(sd_taxable_land, na.rm = TRUE),
    sd_taxable_building = mean(sd_taxable_building, na.rm = TRUE)
  ) 

# Convert the summary data to long format for visualization
library(tidyr)
summary_residential_long <- residential_summary_analysis %>%
  pivot_longer(
    cols = -exemption, 
    names_to = "metric", 
    values_to = "value"
  )
```

```{r}
#Plot Only "avg_market_value" and "sd_market_value"

summary_filtered <- summary_residential_long %>%
  filter(metric %in% c("avg_market_value", "sd_market_value"))

custom_colors <- c("0" = "#E42524",  
                   "1" = "#00ADA9") 

ggplot(summary_filtered, aes(x = metric, y = value, fill = as.factor(exemption), color = as.factor(exemption))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) + 
  geom_text(aes(label = round(value, 0)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 4.5, fontface = "bold") + 
  ylim(0, max(summary_filtered$value) * 1.2) +
  scale_fill_manual(values = custom_colors, labels = c("No Exemption", "With Exemption")) +
  scale_color_manual(values = custom_colors, guide = "none") +  
  labs(title = "Average and Standard Deviation of Market Value (2015-2025)",
       subtitle = "Comparison of Properties With and Without Exemption",
       x = "Metric", 
       y = "Value ($)", 
       fill = "Exemption Status") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 13, color = "gray40"))

```

#### Predictor Variables: Transfers Data

```{r}
transfers2 <- read.csv("Data/RTT_SUMMARY.csv")
```

```{r}
residential_transfers <- residential_properties %>%
  left_join(transfers2, by = c("parcel_number" = "opa_account_num"))
```

```{r}
# Deed Type Analysis
residential_transfers <- residential_transfers %>%
  mutate(document_type = ifelse(is.na(document_type), "Unknown", document_type))
```

```{r}
exemption_by_document_type <- residential_transfers %>%
  group_by(document_type) %>%
  summarise(
    total_count = n(),
    exemption_count = sum(exemption == 1),
    exemption_proportion = exemption_count / total_count * 100
  ) %>%
  arrange(desc(exemption_proportion))

print(exemption_by_document_type)
```

```{r}
## binomial model:DEED - DECEASED, SATISFACTION OF MORTGAGE, MORTGAGE, ASSIGNMENT OF MORTGAGE
document_exemption_model <- glm(exemption ~ document_type, data = residential_transfers, family = binomial)

summary(document_exemption_model)
```

```{r}
# Minimal Financial Consideration Analysis

minimal_threshold <- 10 

minimal_financial_transfers <- residential_transfers %>%
  filter(total_consideration <= minimal_threshold)
```

```{r}

exemption_summary <- minimal_financial_transfers %>%
  group_by(exemption) %>%
  summarise(
    count = n(),
    proportion = count / nrow(minimal_financial_transfers) * 100
  )

print(exemption_summary)
```

```{r}
document_exemption_model <- glm(exemption ~ document_type, data = residential_transfers, family = binomial)

summary(document_exemption_model)
```

```{r}
#Recent(2y) Transfer Analysis
library(dplyr)
library(lubridate)

# Convert recording_date to Date format
residential_transfers <- residential_transfers %>%
  mutate(recording_date = as.Date(recording_date.x, format="%Y-%m-%d"))

# Get the current year
current_year <- year(Sys.Date())
```

```{r}
# Filter and summarize recent transfers
residential_transfers_2y <- residential_transfers %>%
  filter(!is.na(recording_date)) %>%
  group_by(parcel_number) %>%
  summarise(
    latest_transfer_year = max(year(recording_date), na.rm = TRUE),  # Most recent transfer year
    exemption = first(exemption),  # Retain exemption status
    .groups = "drop"
  ) %>%
  mutate(
    latest_transfer_year = ifelse(is.infinite(latest_transfer_year), NA, latest_transfer_year)  # Handle infinite values
  )

# Mark properties with recent transfers (within 2 years)
residential_transfers_2y <- residential_transfers_2y %>%
  mutate(has_recent_transfer = ifelse(!is.na(latest_transfer_year) & latest_transfer_year >= (current_year - 2), 1, 0))

# View results
print(head(residential_transfers_2y))
```

```{r}
# Calculate recent transfer rates by exemption status
exemption_transfer_analysis_residential <- residential_transfers_2y %>%
  group_by(exemption) %>%
  summarise(
    avg_recent_transfer = mean(has_recent_transfer, na.rm = TRUE) * 100  # Convert to percentage
  )

# View results
print(exemption_transfer_analysis_residential)
```

On average, 20.07% of properties without a homestead exemption had a transfer in the past two years, compared to 16.17% of those with an exemption. This indicates that properties without exemptions tend to have a slightly higher likelihood of recent transactions.


```{r}
# Create a contingency table for exemption and recent transfers
exemption_transfer_table <- table(residential_transfers_2y$exemption, residential_transfers_2y$has_recent_transfer)

# Perform chi-square test
chi_test_result <- chisq.test(exemption_transfer_table)

# View test results
print(chi_test_result)
```

#### Dependent Variables: census data

```{r}
# Scatterplot of homestead rates vs median home values
ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), 
       aes(x = median_home_value, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  labs(
    title = "Homestead Exemption Rates vs. Home Values",
    x = "Median Home Value",
    y = "Percentage with Homestead Exemption"
  )

#use liner model =lm
```

```{r}
# Create the linear model
homestead_model <- lm(pct_homestead ~ median_home_value, 
                     data = census_tracts_enriched %>% 
                     filter(pop_density > 0))

# View the summary statistics
summary(homestead_model)

# Visualize with linear fit instead of loess
ggplot(census_tracts_enriched %>% 
       filter(pop_density > 0), 
       aes(x = median_home_value, y = pct_homestead)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +  # Changed from loess to lm
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  labs(
    title = "Linear Relationship: Homestead Exemption Rates vs. Home Values",
    x = "Median Home Value",
    y = "Percentage with Homestead Exemption"
  )

```

The scatterplot shows the relationship between median home values (x-axis) and homestead exemption rates (y-axis) across Philadelphia census tracts. The pattern suggests:

Homestead exemption rates increase with home values up to around $250,000. 

Peak participation occurs in the $200,000-$300,000 range (around 50%).

There's a slight decline in participation for higher-value homes.

The widening gray area at higher home values indicates more uncertainty in the trend, likely due to fewer data points in that range.

Wide variation in participation rates across all home values, shown by the vertical spread of points

```{r}
# Owner Occupancy vs Homestead Rates Analysis
occupancy_analysis <- census_tracts_enriched %>%
  filter(pop_density > 0) %>%
  mutate(
    owner_occ_rate = (owner_hh / occupied_units) * 100,
    pct_homestead_owners = (homestead_count / owner_hh) * 100
  ) %>%
  select(GEOID, owner_occ_rate, pct_homestead_owners)

# Visualize the relationship
ggplot(occupancy_analysis, aes(x = owner_occ_rate, y = pct_homestead_owners)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(
    title = "Owner Occupancy Rate vs. Homestead Participation",
    x = "Owner Occupancy Rate (%)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()

ggsave("outputs/homestead-exemption-distribution.png", census_hist, width = 10, height = 6)
```
This scatter plot reveals important patterns in homestead exemption participation across Philadelphia's neighborhoods. By comparing census tract data on owner occupancy rates (from the 2022 5-year ACS) with homestead exemption enrollment, we can identify areas where participation could be improved.

The data shows that while most Philadelphia census tracts have owner occupancy rates between 25-75%, and generally over half of eligible homeowners participate in the program, there are clear opportunities for improvement. Particularly concerning are:

Census tracts with participation rates below 50%

Areas with high owner occupancy but low program participation

Neighborhoods falling well below the expected participation rate (shown by the blue trend line)

Notably, higher rates of owner occupancy don't automatically translate to higher program participation. This suggests that other factors beyond home ownership - such as awareness of the program, ease of enrollment, or demographic characteristics - may play more significant roles in determining participation rates. These insights can help guide targeted outreach efforts to increase program enrollment among eligible homeowners who are currently missing out on this tax benefit.

```{r}
# Create the dataset with the calculated rates
occupancy_analysis <- census_tracts_enriched %>%
  filter(pop_density > 0) %>%
  mutate(
    owner_occ_rate = (owner_hh / occupied_units) * 100,
    pct_homestead_owners = (homestead_count / owner_hh) * 100
  ) %>%
  select(GEOID, owner_occ_rate, pct_homestead_owners)

# Fit the linear model
occupancy_model <- lm(pct_homestead_owners ~ owner_occ_rate, data = occupancy_analysis)

# View model summary
summary(occupancy_model)

# Visualize with linear fit
ggplot(occupancy_analysis, aes(x = owner_occ_rate, y = pct_homestead_owners)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +  # Changed from loess to lm
  labs(
    title = "Linear Relationship: Owner Occupancy vs. Homestead Participation",
    x = "Owner Occupancy Rate (%)",
    y = "Homestead Participation Rate (%)"
  ) +
  theme_minimal()

```


# Zoning Analysis
```{r}
# Data frame of zoning types
zoning_types <- data.frame(
  Type = c(
    "Single Family Detached",
    "Single Family Attached",
    "Two-Family Attached",
    "Multi-Family",
    "Residential Mixed-Use",
    "Commercial Mixed-Use",
    "Industrial Residential Mixed-Use"
  ),
  Codes = c(
    "RSD1, RSD2, RSD3",
    "RSA1, RSA2, RSA3, RSA4, RSA5, RSA6",
    "RTA1",
    "RM1, RM2, RM3, RM4",
    "RMX1, RMX2, RMX3",
    "CMX1, CMX2, CMX2.5, CMX3, CMX4, CMX5",
    "IRMX"
  ),
  Description = c(
    "Detached houses on individual lots",
    "Attached and semi-detached houses on individual lots",
    "Two-family, semi-detached houses on individual lots",
    "Moderate to high-density multi-unit residential buildings",
    "Residential and mixed-use development, including master plan development",
    "Neighborhood to regional-serving mixed-use development",
    "Mix of low-impact industrial, artisan industrial, residential, and neighborhood commercial uses"
  )
)

# Formatted table
kable(zoning_types,
      col.names = c("Residential Type", "Zoning Codes", "Description"),
      caption = "Philadelphia Residential Zoning Classifications")

```


```{r}
properties_filtered <- filtered_properties %>%
  select(
    zoning,
    homestead_exemption,
    is_residential,
    census_tract,
    shape
  )

# First create a zoning type classification
filtered_properties <- filtered_properties %>% 
  mutate(
    zoning_type = case_when(
      zoning %in% c("RSD1", "RSD2", "RSD3") ~ "Single Family Detached",
      zoning %in% c("RSA1", "RSA2", "RSA3", "RSA4", "RSA5", "RSA6") ~ "Single Family Attached",
      zoning %in% c("RTA1") ~ "Two-Family Attached",
      zoning %in% c("RM1", "RM2", "RM3", "RM4") ~ "Multi-Family",
      zoning %in% c("RMX1", "RMX2", "RMX3") ~ "Residential Mixed-Use",
      zoning %in% c("CMX1", "CMX2", "CMX2.5", "CMX3", "CMX4", "CMX5") ~ "Commercial Mixed-Use",
      zoning %in% c("IRMX") ~ "Industrial Residential Mixed-Use",
      TRUE ~ "Other"
    ),
    is_residential = ifelse(zoning_type != "Other", 1, 0)
  )


zoning_summary <- filtered_properties %>%
  filter(is_residential == 1) %>%
  group_by(zoning_type) %>%
  summarise(
    total_properties = n(),
    homestead_count = sum(homestead_exemption > 0, na.rm = TRUE),
    pct_homestead = (homestead_count / total_properties) * 100
  ) %>%
  arrange(desc(pct_homestead))


# Homestead rates by zoning type
zoning_summary_chart <- ggplot(zoning_summary, 
       aes(x = reorder(zoning_type, -pct_homestead), 
           y = pct_homestead)) +
  geom_bar(stat = "identity", 
           fill = "#e42524",
           alpha = 0.8) +
  geom_text(aes(label = round(pct_homestead,1)), 
            vjust = -0.5,
            size = 3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14)
  ) +
  labs(
    title = "Homestead Exemption Rates by Zoning Category in Philadelphia",
    subtitle = "Residential and Mixed-Use Districts Only",
    x = "Zoning Category",
    y = "Percentage with Homestead Exemption",
    caption = "Source: Philadelphia Property Data, 2025"
  ) +
  scale_y_continuous(
    limits = c(0, max(zoning_summary$pct_homestead) * 1.1),
    labels = function(x) paste0(x, "%")
  )



zoning_summary_chart
ggsave("outputs/homestead-exemption-zoning_summary_chart.png", zoning_summary_chart, width = 10, height = 6)
```

# Modeling



