---
title: "FP-spatial"
author: "Whitney Zhu"
date: "2025-05-08"
output: html_document
---

```{r}
library(readr)
library(dplyr)
fp_data_0505 <- read_csv("~/Downloads/fp_data_0505.csv")
glimpse(fp_data_0505)
```

```{r}
library(sf)

fp_sf_0505 <- st_as_sf(fp_data_0505, coords = c("lon_4326", "lat_4326"), crs = 4326)
```

```{r}
shapefile_path <- "~/Downloads/tl_2020_42101_tract20/tl_2020_42101_tract20.shp"

philly_shapefile <- st_read(shapefile_path)
```

```{r}
library(ggplot2)

ggplot() +
  geom_sf(data = philly_shapefile, fill = "white", color = "black") +  
  geom_sf(data = fp_sf_0505, color = "red", alpha = 0.5, size = 0.1) +       
  labs(
    title = "False Positive Properties in Philadelphia",
    subtitle = "Predicted as Exempted but Actually Not",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```
# Spatial Density of Missed Exemptions
A density map reveals clustering patterns of missed exemptions across the city. Southwest and North Philadelphia emerge as notable hotspots, suggesting spatial barriers or inequities in exemption uptake. These areas may benefit from additional support or communication efforts.
```{r}
ggplot() +
  geom_sf(data = philly_shapefile, fill = "white", color = "gray80") +
  stat_density2d(
    data = fp_data_0505, 
    aes(x = lon_4326, y = lat_4326, fill = ..level..), 
    geom = "polygon", alpha = 0.6, color = NA
  ) +
  scale_fill_viridis_c() +
  coord_sf() +
  labs(title = "Density of False Positive Properties in Philadelphia") +
  theme_minimal()

```

# DBSCAN
```{r}
fp_proj <- st_transform(fp_sf_0505, 26918)

coords <- st_coordinates(fp_proj)
```


```{r}
library(dbscan)

eps_val <- 300   
min_pts <- 20    

db_result <- dbscan(coords, eps = eps_val, minPts = min_pts)

fp_proj$cluster <- as.factor(db_result$cluster)
```

```{r}
ggplot(fp_proj) +
  geom_sf(aes(color = cluster), size = 0.2, alpha = 0.7) +
  scale_color_viridis_d(option = "plasma", name = "Cluster") +
  labs(
    title = "DBSCAN Clustering of False Positive Properties",
    subtitle = paste("ε =", eps_val, ", minPts =", min_pts),
    caption = "Cluster 0 = noise (not assigned to any cluster)"
  ) +
  theme_minimal()
```


# spatial join to Tracts

```{r}

library(dplyr)
library(sf)

philly_shapefile <- st_transform(philly_shapefile, st_crs(fp_proj))
fp_with_tract <- st_join(fp_proj, philly_shapefile)

cluster_summary <- fp_with_tract %>%
  st_drop_geometry() %>%
  group_by(TRACTCE20, cluster) %>%
  summarise(count = n(), .groups = "drop")

print(cluster_summary)
```

```{r}
tract_fp_stats <- cluster_summary %>%
  group_by(TRACTCE20) %>%
  summarise(
    total_fp = sum(count),
    n_clusters = n_distinct(cluster)
  ) %>%
  arrange(desc(total_fp))

head(tract_fp_stats)
```

```{r}
philly_tracts_joined <- left_join(philly_shapefile, tract_fp_stats, by = "TRACTCE20")
```



# Missed Exemption Opportunities by Census Tracts
This map shows the number of false positive predictions—cases where a property was predicted to receive an exemption but does not—in each census tract. Higher counts (in red) indicate areas with greater concentrations of missed exemption opportunities. This spatial breakdown helps target outreach efforts by tract.


```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(scales)

philly_tracts_joined <- left_join(philly_shapefile, tract_fp_stats, by = "TRACTCE20")

top10_tracts <- philly_tracts_joined %>%
  arrange(desc(total_fp)) %>%
  slice_head(n = 10) %>%
  pull(TRACTCE20)

philly_tracts_joined$highlight_top10 <- philly_tracts_joined$TRACTCE20 %in% top10_tracts

ggplot() +
  geom_sf(data = philly_tracts_joined, fill = "white", color = "gray70", size = 0.3) +
  geom_sf(data = filter(philly_tracts_joined, !is.na(total_fp)),
          aes(fill = total_fp), color = NA) +
  
  scale_fill_gradient(low = "#00ADA9", high = "#E42524", name = "False Positive Count") +
  geom_sf(data = filter(philly_tracts_joined, highlight_top10),
          fill = NA, color = "#E42524", linewidth = 0.6) +
  labs(
    title = "Missed Exemption Opportunities by Census Tract",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )
```

```{r}
library(dplyr)

top10_tracts <- tract_fp_stats %>%
  arrange(desc(total_fp)) %>%
  slice(1:10)

print(top10_tracts)

library(knitr)
kable(top10_tracts, caption = "Top 10 Census Tracts by Predicted but Unclaimed Exemptions")

```





# Moran's I
```{r}
install.packages("spdep")  
library(spdep)
```

```{r}
philly_moran <- philly_tracts_joined %>% 
  filter(!is.na(total_fp))

tract_nb <- poly2nb(philly_moran)
tract_weights <- nb2listw(tract_nb, style = "W", zero.policy = TRUE)
```

```{r}
local_moran <- localmoran(
  x = philly_moran$total_fp,
  listw = tract_weights,
  zero.policy = TRUE
)

philly_moran$Ii <- local_moran[, 1]      
philly_moran$Z_Ii <- local_moran[, 4]    
philly_moran$p_value <- local_moran[, 5] 

```

```{r}
fp_mean <- mean(philly_moran$total_fp)

philly_moran <- philly_moran %>%
  mutate(
    lag_fp = lag.listw(tract_weights, total_fp),
    cluster_type = case_when(
      total_fp > fp_mean & lag_fp > fp_mean & p_value < 0.05 ~ "High-High",
      total_fp < fp_mean & lag_fp < fp_mean & p_value < 0.05 ~ "Low-Low",
      total_fp > fp_mean & lag_fp < fp_mean & p_value < 0.05 ~ "High-Low",
      total_fp < fp_mean & lag_fp > fp_mean & p_value < 0.05 ~ "Low-High",
      TRUE ~ "Not Significant"
    )
  )
```

```{r}
library(dplyr)

philly_moran %>%
  count(cluster_type)
```

```{r}
philly_moran %>%
  filter(cluster_type == "High-High") %>%
  summarise(total_fp_in_highhigh = sum(total_fp, na.rm = TRUE))
```

```{r}
library(ggplot2)

ggplot(philly_moran) +
  geom_sf(aes(fill = cluster_type), color = "gray90", size = 0.2) +
  scale_fill_manual(
    values = c(
      "High-High" = "#E42524",
      "Low-Low" = "#00ADA9",
      "High-Low" = "#F39C12",
      "Low-High" = "#9B59B6",
      "Not Significant" = "lightgray"
    ),
    name = "LISA Cluster"
  ) +
  labs(
    title = "LISA Cluster Map (Local Moran's I)",
    subtitle = "Clusters of Missed Exemption Opportunities"
  ) +
  theme_minimal()

```
# Outreach Cost Estimation for High-Potential Clusters
We identified 12,302 properties within statistically significant high-high clusters of missed exemptions using LISA spatial analysis.
This map highlights those target areas (in red) where need is spatially concentrated.

Based on cost estimates from the City’s Opioid Response Unit door-knocking campaign and USPS Direct Mail pricing,
the projected outreach cost for these homes is approximately $84,093.

```{r}
library(ggplot2)

ggplot(philly_moran) +
  geom_sf(aes(fill = cluster_type), color = "gray90", size = 0.2) +
  scale_fill_manual(
    values = c(
      "High-High" = "#E42524"
    ),
    name = "LISA Cluster"
  ) +
  labs(
    title = "LISA Cluster Map (Local Moran's I)",
    subtitle = "Clusters of Missed Exemption Opportunities"
  ) +
  theme_minimal()

```

```{r}
highhigh_tracts <- philly_moran %>%
  filter(cluster_type == "High-High")

highhigh_union <- highhigh_tracts %>%
  st_union() %>%         
  st_sf()               

st_write(highhigh_union, "highhigh_union_0505.geojson", driver = "GeoJSON")
```


```{r}
library(sf)
library(dplyr)

highhigh_sf <- philly_moran %>% 
  filter(cluster_type == "High-High")

highhigh_sf <- highhigh_sf %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(lat = st_coordinates(centroid)[, 2])

lat_mid <- median(highhigh_sf$lat, na.rm = TRUE)

highhigh_summary <- highhigh_sf %>%
  mutate(region = ifelse(lat >= lat_mid, "North Group", "South Group")) %>%
  group_by(region) %>%
  summarise(
    num_tracts = n(),
    total_fp = sum(total_fp, na.rm = TRUE)
  )

print(highhigh_summary)
```



