---
title: "Philadelphia Census Tract Language Analysis"
output: html_document
date: "2025-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

language_map_data <- st_read("Data/properties_without_exemption_language_0505.geojson")
```


```{r}
# Let's look at the structure of the data
str(language_map_data)
```

# Philadelphia Census Tract Language Data

This dataset contains information about the dominant non-English languages spoken across Philadelphia's census tracts, based on the 2020 American Community Survey (ACS) data. The dataset includes 391 census tracts and consists of 5 key variables:

## Dataset Structure
- **Format**: Spatial data frame ('sf' class) with 391 observations
- **Geographic Coverage**: Philadelphia County, Pennsylvania
- **Year**: 2020

## Variables

1. **GEOID** (Character)
   - 11-digit census tract identifier
   - Format: "42101XXXXXX"
   - First 2 digits (42) = Pennsylvania state code
   - Next 3 digits (101) = Philadelphia County code
   - Last 6 digits = Specific tract identifier

2. **NAME** (Character)
   - Simplified census tract names
   - Format: "Tract XX" or "Tract XX.XX"
   - Examples: "Tract 14", "Tract 42.01"

3. **dominant_language** (Character)
   - Most common non-English language in each tract
   - If English is most common, shows second most common language
   - Categories include:
     - Spanish
     - Chinese
     - Vietnamese
     - Russian Polish Slavic
     - Arabic
     - Korean
     - French Haitian Cajun
     - Other Indo European
     - Other Asian Pacific
     - German West Germanic
     - Tagalog
     - Other Unspecified

4. **dominant_pct** (Numeric)
   - Percentage of population speaking the dominant non-English language
   - Values range from approximately 3.7% to 54.2%
   - Rounded to one decimal place

5. **geometry** (sfc_MULTIPOLYGON)
   - Spatial boundaries of each census tract
   - Contains coordinate pairs defining tract shapes
   - Uses XY coordinate system
   - Compatible with mapping software for visualization


```{r}
unique(language_map_data$dominant_language)
```


```{r}
library(ggplot2)
library(sf)
library(dplyr)

language_colors <- c(
  "Spanish" = "#E41A1C",
  "Chinese" = "#377EB8",
  "Vietnamese" = "#4DAF4A",
  "Russian Polish Slavic" = "#984EA3",
  "Arabic" = "#FF7F00",
  "Korean" = "#FFD700",
  "French Haitian Cajun" = "#A65628",
  "Other Indo European" = "#F781BF",
  "Other Asian Pacific" = "#999999",
  "German West Germanic" = "#66C2A5",
  "Tagalog" = "#FC8D62",
  "Other Unspecified" = "#8DA0CB",
  "NA" = "gray90"
)

language_map_data$dominant_language[is.na(language_map_data$dominant_language)] <- "NA"

# 绘制地图
ggplot(language_map_data) +
  geom_sf(aes(fill = dominant_language), color = NA) + 
  scale_fill_manual(values = language_colors) + 
  labs(
    title = "Dominant Non-English Language by Region",
    fill = "Dominant Language"
  ) +
  theme_minimal() + 
  theme(
    legend.position = "bottomright", 
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12) 
  ) #+
  #geom_sf_text(aes(label = sprintf("%s: %g%%", NAME, dominant_pct)), size = 3, color = "black") 
```

```{r}
ggplot(language_map_data) +
  geom_sf(aes(fill = dominant_language), color = "black", size = 0.01) + 
  scale_fill_manual(values = language_colors)
```


```{r cars}
library(leaflet)

# Create the leaflet map with direct color assignment and tract name in label
leaflet(language_map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~case_when(
      dominant_language == "Spanish" ~ "#E41A1C",
      dominant_language == "Chinese" ~ "#377EB8",
      dominant_language == "Vietnamese" ~ "#4DAF4A",
      dominant_language == "Russian Polish Slavic" ~ "#984EA3",
      dominant_language == "Arabic" ~ "#FF7F00",
      dominant_language == "Korean" ~ "#FFD700",
      dominant_language == "French Haitian Cajun" ~ "#A65628",
      dominant_language == "Other Indo European" ~ "#F781BF",
      dominant_language == "Other Asian Pacific" ~ "#999999",
      dominant_language == "German West Germanic" ~ "#66C2A5",
      dominant_language == "Tagalog" ~ "#FC8D62",
      dominant_language == "Other Unspecified" ~ "#8DA0CB",
      TRUE ~ "#FFFFFF"  # Default white for any unmatched categories
    ),
    weight = 0.5,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~sprintf(
      "%s<br>%s: %g%%",
      NAME,
      dominant_language,
      dominant_pct
    ) %>% lapply(htmltools::HTML),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", 
               "#FF7F00", "#FFD700", "#A65628", "#F781BF",
               "#999999", "#66C2A5", "#FC8D62", "#8DA0CB"),
    labels = c("Spanish", "Chinese", "Vietnamese", "Russian Polish Slavic",
              "Arabic", "Korean", "French Haitian Cajun", "Other Indo European",
              "Other Asian Pacific", "German West Germanic", "Tagalog", "Other Unspecified"),
    title = "Dominant Non-English Language",
    opacity = 0.7
  )
```

Here's an explanation of what this map shows:

This interactive map visualizes the dominant non-English languages spoken across Philadelphia's census tracts based on 2020 American Community Survey (ACS) data. For each census tract, the map displays:

1. **Spatial Distribution**: Each colored polygon represents a census tract in Philadelphia, with colors indicating the most common non-English language spoken in that tract (after English).

2. **Color Coding**:
- Red areas indicate Spanish as the dominant non-English language
- Blue represents Chinese
- Green shows Vietnamese
- Purple indicates Russian/Polish/Slavic languages
- Orange represents Arabic
- Yellow shows Korean
- Brown indicates French/Haitian/Cajun
- Pink represents Other Indo-European languages
- Grey shows Other Asian/Pacific languages
- Teal represents German/West Germanic languages
- Coral indicates Tagalog
- Light blue shows Other/Unspecified languages

3. **Interactive Features**:
- Hovering over any tract reveals:
  - The census tract name/number
  - The dominant non-English language in that tract
  - The percentage of residents who speak that language

This visualization helps identify linguistic communities throughout Philadelphia and can be valuable for:
- Planning language access services
- Targeting community outreach
- Understanding neighborhood demographics
- Identifying areas where translation services might be needed
- Supporting cultural and educational programming decisions

The map reveals clear patterns of language communities in Philadelphia, such as Spanish-speaking concentrations in certain neighborhoods, Asian language clusters in others, and the overall linguistic diversity of the city.



# Join w/ Predictions
```{r pressure, echo=FALSE}
# First read in both datasets
language_data <- st_read("./Data/philly_language_dominance.geojson")

properties_no_exemption <- st_read("./data/properties_without_exemption_language_0505.geojson")

# Join the datasets, keeping only the new columns we want from language_data
properties_with_language <- properties_no_exemption %>%
  left_join(
    language_data %>%
      st_drop_geometry() %>%  # Drop geometry since we'll keep the geometry from properties
      select(GEOID, NAME, dominant_language, dominant_pct) %>%
      rename(tract_name = NAME),  # Rename to avoid any confusion with other NAME columns
    by = "GEOID"
  )

# Save the new combined dataset
#st_write(properties_with_language, 
         #"./data/properties_without_exemption_language_0505.geojson", 
         #delete_dsn = TRUE)
```


```{r}
names(properties_with_language)

# Look at a few rows to see how the new data looks
head(properties_with_language %>% 
     select(GEOID, tract_name, dominant_language, dominant_pct))
```

I have updated property_without_exemption_tractdata.geojson and uploaded a new file called properties_without_exemption_language.geojson.

In it there are three new columns

1. tract_name
The census tract identifier in a readable format
Example: "Tract 14" or "Tract 42.01"

2. dominant_language
The most common non-English language in that census tract
Values include: "Spanish", "Chinese", "Vietnamese", etc.
If English is the most common language in the tract, this shows the second most common language

3.dominant_pct
The percentage of people who speak the dominant non-English language in that tract
Numeric value rounded to one decimal place
Example: 5.7 means 5.7% of people in that tract speak the dominant non-English language

```{r}
# Summary of dominant languages across all tracts
language_summary <- language_data %>%
  st_drop_geometry() %>%
  group_by(dominant_language) %>%
  summarise(
    number_of_tracts = n(),
    avg_percentage = mean(dominant_pct, na.rm = TRUE),
    min_percentage = min(dominant_pct, na.rm = TRUE),
    max_percentage = max(dominant_pct, na.rm = TRUE)
  ) %>%
  arrange(desc(number_of_tracts)) %>%
  mutate(
    pct_of_tracts = (number_of_tracts / sum(number_of_tracts) * 100)
  ) %>%
  knitr::kable(
    col.names = c(
      "Dominant Language",
      "Number of Tracts",
      "Average %",
      "Minimum %",
      "Maximum %",
      "% of All Tracts"
    ),
    digits = 1,
    caption = "Distribution of Dominant Non-English Languages Across Philadelphia Census Tracts"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    font_size = 18,
    full_width = TRUE
  ) %>%
  kableExtra::row_spec(0, bold = TRUE, font_size = 19)  # Makes header larger

# Top 10 tracts with highest language percentages
top_tracts <- language_data %>%
  st_drop_geometry() %>%
  arrange(desc(dominant_pct)) %>%
  head(10) %>%
  select(NAME, dominant_language, dominant_pct) %>%
  knitr::kable(
    col.names = c(
      "Census Tract",
      "Dominant Language",
      "Percentage"
    ),
    digits = 1,
    caption = "Top 10 Census Tracts with Highest Non-English Language Concentrations"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    font_size = 18,
    full_width = TRUE
  ) %>%
  kableExtra::row_spec(0, bold = TRUE, font_size = 19)  # Makes header larger

# Display both tables
language_summary
top_tracts
```

```{r}
# DO NOT RUN THIS CODE- this is from the first join before updating predictions. This is for comparison only
# Check the new columns
names(properties_with_language)

# Look at a few rows to see how the new data looks
head(properties_with_language %>% 
     select(GEOID, tract_name, dominant_language, dominant_pct))

[1] "OBJECTID"          "Join_Count"        "TARGET_FID"        "objectd"           "cnss_tr"           "exemptn"           "X_pred1"          
 [8] "exm_0_5"           "exm_0_6"           "exm_0_7"           "exm_0_8"           "exm_0_9"           "sm_ddrs"           "is_deep"          
[15] "large_r"           "ownr_cn"           "rntl_lc"           "cmmrcl_"           "blnc_rt"           "avg_mr_"           "sd_mrk_"          
[22] "hs_rcn_"           "ltst_d_"           "pct_fr_"           "ovrll__"           "pvrty_r"           "bch_dg_"           "yng_wn_"          
[29] "snr_wn_"           "fmly_h_"           "lmtd_n_"           "dvrsty_"           "mdn_ncm"           "ownr_c_"           "cst_br_"          
[36] "mrtgg__"           "mdn_hm_"           "GEOID"             "ALAND20"           "medin_g"           "totl_pp"           "occpd_n"          
[43] "vcnt_nt"           "rntr_hh"           "ownr_hh"           "pct_wht"           "pct_blc"           "pct_hsp"           "pct_asn"          
[50] "pct_fr1"           "ovrll_1"           "ownr_v_"           "rntl_v_"           "bch_dg1"           "pvrty_r_1"         "yng_wn1"          
[57] "mddl_w_"           "snr_wn1"           "fmly_h1"           "lmtd_n1"           "pp_dnst"           "mdn_ncm_1"         "mdn_hm1"          
[64] "mdn_mlt"           "cst_br1"           "mrtgg_1"           "dvrsty1"           "gntrfc_"           "tract_name"        "dominant_language"
[71] "dominant_pct"      "geometry"         
Simple feature collection with 6 features and 4 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: -75.25724 ymin: 39.9228 xmax: -75.12713 ymax: 40.05616
Geodetic CRS:  WGS 84
        GEOID   tract_name   dominant_language dominant_pct                   geometry
1 42101006400     Tract 64   Other Unspecified         11.2  POINT (-75.24368 39.9228)
2 42101015102 Tract 151.02             Spanish          0.6 POINT (-75.17936 39.98911)
3 42101016500    Tract 165             Spanish         11.1 POINT (-75.14811 39.99009)
4 42101022000    Tract 220 Other Indo European          2.6 POINT (-75.25724 40.05616)
5 42101002400     Tract 24             Spanish          7.3 POINT (-75.15825 39.93792)
6 42101016002 Tract 160.02             Spanish          3.9 POINT (-75.12713 39.97713)
```

